---
title: "4nqo+e7386 Global Analysis"
author: "M. Muzamil Khan"
date: '2022-09-14'
output:
  html_document:
    code_folding: hide
    theme: flatly
    toc: yes
    toc_float: true
  html_notebook:
    code_folding: hide
    theme: flatly
    toc: yes
    toc_float: true
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
.libPaths("/restricted/projectnb/montilab-p/personal/mmkhan/R/4.1/")
PATH <- "/restricted/projectnb/montilab-p/projects/oralcancer/4nqo/"
MLAB <- "/restricted/projectnb/montilab-p/"
library(Seurat)
library(ggplot2)
library(dplyr)
library(gtsummary)
library(gt)
library(webshot)
library(hypeR)
library(stringr)
```

# Analysis of e7386-related samples.

This contains samples from ctrl, 4nqo-ctrl, e7386@{25,50} mg/kg with two replicates each. This file contains:
1. Creating seurat object from relevant samples.
2. Normalisation, clustering and dimensionality reduction.
3. singleR cell-type annotation (not run here but stratification done per cell)
4. cluster and trt group level differential and pathway analysis.
5. Misc analysis of cell-type exploration and sigs stratification.

## Create Seurat object
### Load data
Extract relevant samples and create a subset

```{r eval=FALSE}
e7386_obj <- readRDS(file.path(PATH, 'lina/data/4nqo_filtered_01312022.rds'))
e7386_obj <- subset(e7386_obj, treatment %in% c('control', '4NQO control', '4NQO + E7386; 25 mg/kg', '4NQO+ E7386; 50 mg/kg'))
e7386_obj$treatment <- factor(e7386_obj$treatment, levels=c('control', '4NQO control', '4NQO + E7386; 25 mg/kg', '4NQO+ E7386; 50 mg/kg'))
```

## Simple QC plots 

```{r eval=FALSE, fig.width=8, fig.height=7}
e7386_obj <- PercentageFeatureSet(e7386_obj, pattern = "^MT-", col.name = "percent.mt")
# Visualize QC metrics as a violin plot
VlnPlot(e7386_obj, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size = 0.0001, group.by = "treatment")
DimPlot(e7386_obj, reduction = "umap", group.by = "batches")
DimPlot(e7386_obj, reduction = "umap", group.by = "Phase")
```

## Normalization and Clustering
Note: for cells >50K, `FindClusters()` throws an error: `Error in asMethod(object) : Cholmod error 'problem too large' at file ../Core/cholmod_dense.c, line 102` that comes from leidenalg in order to resolve this set `method="igraph"`. More details [here](https://github.com/satijalab/seurat/issues/2294)
The elbow is formed at PCs 12-15 hence 1:15 PCs are chosen for downstream analyses. The resolution parameter at 0.3 gives 17 clusters, change to 0.5 (=22 clusters) to get ample clusters to work with.

```{r eval=F, eval=FALSE}
e7386_obj <- NormalizeData(e7386_obj, normalization.method = "LogNormalize", scale.factor = 10000)
e7386_obj <- FindVariableFeatures(e7386_obj, selection.method = "vst", nfeatures = 5000)
all.genes <- rownames(e7386_obj)
e7386_obj <- ScaleData(e7386_obj, all.genes)

# run sctransform
e7386_obj <- SCTransform(e7386_obj, vars.to.regress = "percent.mt", verbose = FALSE)
e7386_obj <- RunPCA(e7386_obj, verbose = FALSE, )

e7386_obj <- FindNeighbors(e7386_obj, dims = 1:15, verbose = FALSE, )
e7386_obj <- FindClusters(e7386_obj, verbose = FALSE, resolution = 0.5, random.seed = 1234, algorithm = "leiden", method = "igraph")
DimPlot(e7386_obj, label = TRUE) + NoLegend()
e7386_obj <- RunUMAP(e7386_obj, dims = 1:15, verbose = FALSE)

#saveRDS(e7386_obj, file.path(PATH, "mmkhan_e7386/data/2023_01_09_e7386_seurat_filtered_main_final.rds"))
```

```{r}
e7386_obj <- readRDS(file.path(PATH, "mmkhan_e7386/data/2023_01_09_e7386_seurat_filtered_main_final.rds"))
e7386_obj$cell_id <- rownames(e7386_obj@meta.data)
DimPlot(e7386_obj, reduction = "umap", group.by = "Phase")
```

Explained PC variance

```{r fig.width=12}
ElbowPlot(object = e7386_obj, reduction = "pca")
```


```{r fig.width=12}
mat <- Seurat::GetAssayData(e7386_obj, assay = "RNA", slot = "scale.data")
pca <- e7386_obj[["pca"]]

# Get the total variance:
total_variance <- sum(matrixStats::rowVars(mat))

eigValues = (pca@stdev)^2  ## EigenValues
varExplained = eigValues / total_variance
varExplained
sum(varExplained[1:15])

```

```{r}
d1 <- DimPlot(e7386_obj, reduction = "umap", label = T, group.by = "seurat_clusters") + NoLegend()
d1

d2 <- DimPlot(e7386_obj, reduction = "umap", label = T, group.by = "seurat_clusters", split.by = "treatment") + NoLegend()
d2

d3 <- DimPlot(e7386_obj, reduction = "umap", label = T, group.by = "treatment") 
d3

d4 <- DimPlot(e7386_obj, reduction = "umap", label = T, split.by = "treatment") 
d4
#ggsave(plot = d1, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/umap_seurat_clusters.png"), width = 5, height = 5)
#ggsave(plot = d2, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/umap_seurat_clusters_split.png"), width = 14, height = 5)
```


## Cluster Analysis

### Silhouette score
calculates distance of cells within a cluster and the nearest cluster => shows similarity with other clusters.

```{r}
library(cluster, quietly = TRUE)
dist.matrix <- dist(x = Embeddings(object = e7386_obj[["pca"]])[, 1:15])
clusters <- e7386_obj$seurat_clusters
sil <- silhouette(x = as.numeric(x = as.factor(x = clusters)), dist = dist.matrix)
e7386_obj$sil <- sil[, 3]
```


```{r}
library(bluster)
library(SingleCellExperiment)
sce <- as.SingleCellExperiment(e7386_obj)
sil.approx <- approxSilhouette(reducedDim(sce, "PCA"), clusters=sce$seurat_clusters)
sil.approx

sil.data <- as.data.frame(sil.approx)
sil.data$closest <- factor(ifelse(sil.data$width > 0, sce$seurat_clusters, sil.data$other))
sil.data$cluster <- sce$seurat_clusters

library(ggplot2)
ggplot(sil.data, aes(x=cluster, y=width, colour=closest)) +
    ggbeeswarm::geom_quasirandom(method="smiley")

```

```{r}
table(Cluster=sce$seurat_clusters, sil.data$closest)
```

### Relative Proportion

```{r ftest_wrapper}
ftest <- function(obj, samples, cond1, cond2){
  df1 <- obj@meta.data
  df1 <- df1[df1$treatment %in% samples,]
  df1$treatment <- factor(df1$treatment, levels = samples)
  tmp <- table(df1[,cond1], df1[,cond2])
  df2 <- data.frame(tmp[,1], tmp[,2])
  colnames(df2) <- colnames(tmp)
  df2 <- df2 %>%
    dplyr::mutate(pval=sapply(
    rownames(df2), function(z) fisher.test(table(df1[,cond1]==z,df1[,cond2]))$p.value)) %>%
      dplyr::mutate(qval=p.adjust(pval, method = "BH")) %>%
      dplyr::mutate(sig=ifelse(qval<=0.01, "***",  ifelse(qval <= 0.05, "*", ""))) %>%
      dplyr::arrange(pval)
  df2

}

```

```{r relative_proportion_wrapper}
rel_prop <- function(obj, cond1, cond2, cols=NULL) {
  meta_data <- obj@meta.data
  dat <- as.data.frame(prop.table(table(meta_data[,cond1], meta_data[,cond2]), margin = 1))
  colnames(dat) <- c(cond1, cond2, "percent")
  
  dat <- dat %>% 
  group_by_at(cond1) %>% 
  mutate(cluster_label_y = 1 - (cumsum(percent) - 0.5 * percent)) %>% 
  ungroup()
  
  d1 <- ggplot(dat, aes_string(cond1, y = "percent", fill = cond2)) +
  geom_col(position = "fill", stat = "identity", width = 0.5) +
  scale_y_continuous(labels = scales::percent) +
  geom_text(aes(y = cluster_label_y, label = round(100 * percent, 2)), col="black", fontface="bold", size=2)+
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, margin=ggplot2::margin(5, b=10)), text = element_text(family = "Arial"))
  
 # d1 <- ggplot(dat, aes(treatment, y = percent, fill = cell_type)) +
 #  #geom_bar(stat = "identity") +
 #  geom_col(position="fill", width=0.5)+
 #  #theme(aspect.ratio = 3)+
 #  scale_fill_manual(values = c("tomato", "pink",  "green3", "turquoise3", "black", "peru"))+
 #  scale_y_continuous(labels = scales::percent) +
 #  geom_text(aes(y = cluster_label_y, label = round(100 * percent, 2)), size=2, col="black")+ 
 #  theme_bw() +   ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))

  if(!is.null(cols)){ 
    d <- d1 +  scale_fill_manual(values = cols)
  } else {
    d <- d1
  }
  d
}


```

Clusters 1,4,7,22 are more of control cells and clusters 9,12,13,15,17,21 are more of treated groups.

```{r}
rel_prop(obj = e7386_obj, cond1 = "treatment", cond2 = "seurat_clusters")

r2 <- rel_prop(obj = e7386_obj, cond2 = "treatment", cond1 = "seurat_clusters", cols = c("gray", "orangered", "yellow", "chocolate"))
r2
#ggsave(plot = r2, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/clust_samples.png"), width = 12, height = 5, dpi = 300)
```


```{r}
dat <- as.data.frame(prop.table(table(e7386_obj$seurat_clusters, e7386_obj$Sample_ID), margin = 1))
colnames(dat) <- c("cluster", "sample", "percent")
dat$sample <- factor(dat$sample, levels = c("KM_BM_0318_1C",  "KM_BM_0319_1", "KM_BM_0318_5NC", "KM_BM_0319_5", "KM_BM_0318_6E", "KM_BM_0319_6", "KM_BM_0318_7E", "KM_BM_0319_7"))

ord <- c("KM_BM_0318_1C",  "KM_BM_0319_1", "KM_BM_0318_5NC", "KM_BM_0319_5", "KM_BM_0318_6E", "KM_BM_0319_6", "KM_BM_0318_7E", "KM_BM_0319_7")
dat <- dat[order(match(dat$sample, ord)),]
dat <- dat %>% 
  group_by(cluster) %>% 
  mutate(cluster_label_y = 1 - (cumsum(percent) - 0.5 * percent)) %>% 
  ungroup()

d3 <- ggplot(dat, aes(cluster, y = percent, fill = sample)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_fill_manual(values = c("gray70", "gray60", "orangered4", "orangered3", "yellow2", "yellow3", "chocolate1", "chocolate3"))+
  scale_y_continuous(labels = scales::percent) +
  geom_text(aes(y = cluster_label_y, label = round(100 * percent, 2)), col="white", fontface="bold", size=2.5)+theme_bw()
d3

#ggsave(plot = d3, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/samplewise_seurat_clusters_prop.png"), width = 12, height = 5)
```

## Summary of cells/cluster/group

```{r fig.width=12}

dat_tab <- e7386_obj@meta.data[, c('treatment', 'seurat_clusters')]
#colnames(dat_tab) <- c("cluster", "treatment", "freq")

tbl_summary_1 <- tbl_summary(dat_tab, by = seurat_clusters)

gt <- tbl_summary_1  %>%  as_gt()   # convert to gt table

gt 

# gt %>% tab_options(column_labels.font.size=10, table.font.size = 8) %>% gt::gtsave(# save table as image
#     filename = "../mmkhan_e7386/results/e7386_images/clust_tab.png",
#   ) 
# 
# webshot("../mmkhan_e7386/results/e7386_images/clust_tab.png", "../mmkhan_e7386/results/e7386_images/clust_tab_adj.png", vwidth  = 1300, vheight = 500)
```

## Cell-type Annotation

### SingleR
```{r eval=F}
#immgen ref sce
ig <- SingleR::ImmGenData()
rownames(ig) <- toupper(rownames(ig))
labels <- ig$label.main

singler_e7386 <- as.SingleCellExperiment(e7386_obj)
logcounts <- LogNormalize(assay(singler_e7386))
assay(singler_e7386, "logcounts") <- logcounts
assay(singler_e7386, "counts") <- NULL

predicted <- SingleR(test = singler_e7386, ref = ig, assay.type.test = 1,labels = labels)
e7386_obj$singleRident <- predicted$labels[match(e7386_obj$cell_id, rownames(predicted))]
e7386_obj$singleRcelltype <- predicted$pruned.labels[match(e7386_obj$cell_id, rownames(predicted))]
saveRDS(e7386_obj, file.path(PATH, "mmkhan_e7386/data/2023_01_09_e7386_seurat_filtered_main_final.rds"))
```


```{r fig.width=12}
`%ni%` <- Negate(`%in%`)
e7386_obj_sub <- subset(e7386_obj, singleRcelltype %ni% c(NA, 'Basophils', 'Microglia'))

#combine all B cells into one
e7386_obj_sub$singleRcelltype <- ifelse(e7386_obj_sub$singleRcelltype %in% c('B cells', 'B cells, pro'), "B cells", e7386_obj_sub$singleRcelltype)
#combine all NK cells into one
e7386_obj_sub$singleRcelltype <- ifelse(e7386_obj_sub$singleRcelltype %in% c('NK cells', 'NKT'), "NK cells", e7386_obj_sub$singleRcelltype)

d4 <- DimPlot(e7386_obj_sub, reduction = "umap", label = T, label.box =F, group.by = 'singleRcelltype', label.size = 3.5)
d4
#ggsave(plot = d4, filename = file.path(PATH, "results/4nqo_e7386_images/singleR_all.png"), width = 9, height = 5)

d5 <- DimPlot(e7386_obj_sub, reduction = "umap", split.by = "treatment", label = T, group.by = 'singleRcelltype',  label.size = 3.5, shuffle = T) 
d5
#ggsave(plot = d5, filename = file.path(PATH, "results/4nqo_e7386_images/singleR_all_w_labels.png"), width = 13, height = 5)

e7386_obj_sub$cell_type <- ifelse(e7386_obj_sub$singleRcelltype %in% c("Epithelial cells", "Endothelial cells", "Stromal cells", "Fibroblasts", "Stem cells"), e7386_obj_sub$singleRcelltype, "Immune")
d6 <- DimPlot(e7386_obj_sub, reduction = "umap",  split.by = "treatment",  label = TRUE, group.by = "cell_type", cols = c("tomato", "pink", "green3", "turquoise3", "black", "peru"), label.size = 5) + NoLegend()
d6
#ggsave(plot = d6, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/umap_coarse_cell_types.png"), width = 13, height = 5)

d6 <- DimPlot(e7386_obj_sub, group.by = "cell_type", reduction = "umap",  label = T, cols = c("tomato", "pink", "green3", "turquoise3", "black", "peru"), label.size = 5) + NoLegend()
d6
#ggsave(plot = d6, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/umap_coarse_cell_types_nolabel.png"), width = 6, height = 5)
```

### Table Summary

```{r}
dat_tab1 <- e7386_obj_sub@meta.data[, c('treatment', 'cell_type')]
tbl_summary_2 <- tbl_summary(dat_tab1, by = cell_type, percent = "row")

gt2 <- tbl_summary_2  %>%  as_gt()   # convert to gt table
gt2
# gt2 %>% tab_options(column_labels.font.size=10, table.font.size = 10) %>% gt::gtsave(# save table as image
#     filename = "../mmkhan_e7386/results/e7386_images/clust_samp_tab.png",
#   ) 
# 
# webshot("../mmkhan_e7386/results/e7386_images/clust_samp_tab.png", "../mmkhan_e7386/results/e7386_images/clust_samp_tab_adj.png", vwidth  = 1000, vheight = 500)

dat_tab1 <- e7386_obj_sub@meta.data[, c('treatment', 'singleRcelltype')]
tbl_summary_2 <- tbl_summary(dat_tab1, by = singleRcelltype, percent = "row")

gt2 <- tbl_summary_2  %>%  as_gt()   # convert to gt table
gt2
# gt2 %>% tab_options(column_labels.font.size=10, table.font.size = 8) %>% gt::gtsave(# save table as image
#     filename = "../mmkhan_e7386/results/e7386_images/trt_singleR.png",
#   ) 
# 
# webshot("../mmkhan_e7386/results/e7386_images/trt_singleR.png", "../mmkhan_e7386/results/e7386_images/trt_singleR.png", vwidth  = 1000, vheight = 500)

```

### Quantify and test significance

```{r}
r1 <- rel_prop(obj = e7386_obj_sub, cond1 = "treatment", cond2 = "cell_type", cols = c("tomato", "pink", "green3", "turquoise3", "black", "peru"))
r1
#ggsave(plot = r1, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/coarse_cell_type_prop.png"), width = 4, height = 5, dpi = 300)

rel_prop(obj = e7386_obj_sub, cond1 = "seurat_clusters", cond2 = "cell_type", cols = c("tomato", "pink", "green3", "turquoise3", "black", "peru"))
```


```{r gene_expression_boxplot_wrapper}
ge_boxplot <- function(genes, obj, cond, ncol){
  df <- data.frame(t(obj@assays$RNA@scale.data[c(genes),]))
  condition <- cond
  #cond <- apply(obj[, cond, drop = FALSE], 1, paste, collapse = "_")
  df[, paste(cond)] <-  apply(obj@meta.data[, cond, drop = FALSE],1, paste, collapse = "_")
  df[, paste(cond)] <- factor(e7386_obj$treatment, levels = c('control', '4NQO control', '4NQO + E7386; 25 mg/kg', '4NQO+ E7386; 50 mg/kg'))

  gsvaResFlat <- reshape2::melt(df, id.vars = paste(cond, collapse = "_"), variable.name = "group")

  g1 <- ggplot2::ggplot(gsvaResFlat, ggplot2::aes_string(
    x = paste(condition, collapse = "_"), y = "value",
    color = paste(condition, collapse = "_"))) +
    ggplot2::geom_boxplot(outlier.shape = NA) +
    #ggplot2::geom_jitter(size=0.0) +
    ggplot2::facet_wrap(~group, scale = "free_y",
                        ncol = ncol) +
    ggpubr::stat_compare_means(method = "anova",  label = 'p.format')+
    ggplot2::theme(strip.text.x = ggplot2::element_text(size = 10))+
    ggplot2::theme(axis.text.x = element_text(angle = 45, hjust = 1))+
    labs(title = 'Gene Expression', y = "Counts", x = cond)+
    theme(axis.text = element_text(size = 10), axis.title = element_text(size = 10, face = "bold"))
  g1
}

```


### FCM 

Stringent cut-off

Remove clusters 2 and 5 due to high mt content and low gene features

```{r}
VlnPlot(e7386_obj, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), pt.size = 0.0001, stack = T)
```

FindConservedMarkers() from Seurat
New cluster labels based on FCM

```{r}
csv <- read.csv("/restricted/projectnb/montilab-p/personal/eric/e7386_eric/results/tables/cluster_markers_allCells.csv")
#write.csv(x = csv, file = file.path(PATH, "mmkhan_e7386/results/2024_02_05_fcm_all_eric.csv"), quote = F)

fcm <- read.table(file = file.path(MLAB, "personal/eric/e7386_eric/data/reference/celltype_all.txt"), header = T, sep = "\t")
tibble(fcm)  %>% print(n = 20)
```



```{r}
`%ni%` <- Negate(`%in%`)
Idents(e7386_obj) <- "seurat_clusters"
fcm_cell_type <- c("Endothelial", "Epithelial", "Fibroblast", "Epithelial", "Endothelial", "Fibroblast", "Epithelial", "Epithelial", "Epithelial", "Epithelial", "Epithelial", "Immune", "Immune", "Glial", "Erythroid", "Epithelial", "Immune", "Muscle", "Endothelial", "Epithelial", "Muscle", "Adipose")
levels(Idents(e7386_obj)) <- fcm_cell_type
e7386_obj$fcm_cell_type <- Idents(e7386_obj)

Idents(e7386_obj) <- "seurat_clusters"
fcm_cell_label <- c("Endothelial", "Epithelial", "Fibroblast", "Basal Epi.", "Endothelial", "Fibroblast", "Suprabasal Epi.", "Proliferating Epi.", "Basal Epi.", "Basal Epi.", "Acinar SLG", "Immune", "Immune", "Glial", "Erythroid", "Serous Acinar SLG", "Immune", "SMC", "LEC", "Acinar SLG", "Chondrocyte-like SMC", "Adipocytes")
levels(Idents(e7386_obj)) <- fcm_cell_label
e7386_obj$fcm_cell_label <- Idents(e7386_obj)

Idents(e7386_obj) <- "seurat_clusters"
e7386_obj <- subset(e7386_obj, seurat_clusters %ni% c(2,5) & singleRcelltype %ni% c(NA, 'Basophils', 'Microglia'))
table(e7386_obj$singleRcelltype)
table(e7386_obj$seurat_clusters)
```

Rename gene names with title case

```{r}
RenameGenesSeurat <- function(obj, RNA_newnames_list, SCT_newnames_list) { # Replace gene names in different slots of a Seurat object. Run this before integration. Run this before integration. It only changes obj@assays$RNA@counts, @data and @scale.data.
  print("Run this before integration. It only changes obj@assays$RNA@counts, @data and @scale.data.")
  RNA <- obj@assays$RNA
  SCT <- obj@assays$SCT
  
  RNA@counts@Dimnames[[1]] <- RNA_newnames_list[[1]]
  RNA@data@Dimnames[[1]] <- RNA_newnames_list[[2]]
  rownames(RNA@scale.data) <- RNA_newnames_list[[3]]
  
  SCT@counts@Dimnames[[1]] <- SCT_newnames_list[[1]]
  SCT@data@Dimnames[[1]] <- SCT_newnames_list[[2]]
  rownames(SCT@scale.data) <- SCT_newnames_list[[3]]
  
  obj@assays$RNA <- RNA
  obj@assays$SCT <- SCT
  
  return(obj)
}

RNA_newnames_list <- list(counts=str_to_title(rownames(e7386_obj@assays$RNA@counts)), data=str_to_title(rownames(e7386_obj@assays$RNA@data)), scale.data=str_to_title(rownames(e7386_obj@assays$RNA@scale.data)))
SCT_newnames_list <- list(counts=str_to_title(rownames(e7386_obj@assays$SCT@counts)), data=str_to_title(rownames(e7386_obj@assays$SCT@data)), scale.data=str_to_title(rownames(e7386_obj@assays$SCT@scale.data)))

new_obj <- RenameGenesSeurat(obj = e7386_obj, RNA_newnames_list = RNA_newnames_list, SCT_newnames_list = SCT_newnames_list)
```

### Cellular compartmental markers

```{r fig.width=10, fig.height=8}

new_obj$new_fcm_cell_type <- ifelse(new_obj$fcm_cell_type %in% c('Glial', 'Erythroid', 'Muscle', 'Adipose'), as.character('Others'), as.character(e7386_obj$fcm_cell_type))
new_obj$new_fcm_cell_type <- factor(new_obj$new_fcm_cell_type)

cols_ct <- rev(scales::hue_pal()(5))
d <- DotPlot(object = new_obj, features = stringr::str_to_title(c("PECAM1", "S1PR1", "CYYR1", "CCL21A", "LYVE1", "CDH1", "EPCAM", "KRT5", "KRT15", "KRT4", "KRT13", "MKI67", "AQP5",  "DPT", "COL1A1", "MGP", "S100A8", "S100A9", "CD74", "PTPRC",
                     "MPZ", "PLP1", "ACTA2", "TAGLN", "CYTL1", "CHAD", "ALAS2")), group.by = "new_fcm_cell_type", assay = "RNA", scale = T, cols = c("white", "purple")) +  theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, size = 15, family = "Arial", face = "bold",), axis.text.y = ggplot2::element_text(size = 13, family = "Arial", face = "bold", colour= cols_ct), panel.grid = element_blank(), 
                 axis.title.x = ggplot2::element_text(size = 15, family = "Arial", face = "bold"),  axis.title.y = ggplot2::element_text(size = 10, family = "Arial", face = "bold"), 
                 legend.text =  ggplot2::element_text(size = 15, family = "Arial", face = "bold"),
                 legend.title = ggplot2::element_text(size = 15, family = "Arial", face = "bold"),
                 axis.title.y.left = ggplot2::element_text(size = 13, family = "Arial", face = "bold"))+ylab("Cellular Compartments")+
  scale_y_discrete(limits=rev) 
d
#ggsave(plot = d, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/dotplot_markers.png"), width = 14, height = 3, dpi = 200)


s <- Seurat::DoHeatmap(object = new_obj, group.bar = T,
        features = str_to_title(c("PECAM1", "S1PR1", "CYYR1", "CCL21A", "LYVE1", "CDH1", "EPCAM", "KRT5", "KRT15", "KRT4", "KRT13", "MKI67", "AQP5", "SMGC", "MUC5B", "DPT", "COL1A1", "MGP", "S100A8", "S100A9", "CD74", "PTPRC",
                     "MPZ", "PLP1", "ACTA2", "TAGLN", "CYTL1", "CHAD", "ALAS2")),  group.by= "new_fcm_cell_type", assay = "SCT", slot = "scale.data") + scale_fill_gradientn(colors = c("pink", "white", "purple"))+
  ggplot2::theme(axis.text.y = ggplot2::element_text(size = 15, family = "Arial", face = "bold"), panel.grid = element_blank(), 
                 axis.text.x = ggplot2::element_text(size = 15, family = "Arial", face = "bold"),  
                 legend.text =  ggplot2::element_text(size = 15, family = "Arial", face = "bold"),
                 legend.title = ggplot2::element_text(size = 15, family = "Arial", face = "bold"))
#ggsave(plot = s, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/heatmap_markers.png"), width = 13, height = 10, dpi = 200)
```



```{r}
d1 <- DimPlot(e7386_obj, reduction = "umap", label = T, split.by = "treatment") + NoLegend()
d1
#ggsave(plot = d1, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/01_09_umap_seurat_clusters_trt.png"), width = 12, height = 5)

d2 <- DimPlot(e7386_obj, reduction = "umap", label = T, group.by = "fcm_cell_type") + NoLegend()
d2
#ggsave(plot = d2, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/01_09_umap_fcm_cell_type.png"), width = 5, height = 5)

d3 <- DimPlot(e7386_obj, reduction = "umap", label = T, group.by = "fcm_cell_label") + NoLegend()
d3
#ggsave(plot = d3, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/01_09_umap_fcm_cell_label.png"), width = 5, height = 5)

d4 <- DimPlot(e7386_obj, reduction = "umap", label = T, group.by = "fcm_cell_type", label.size = 5) + 
  ggplot2::theme(strip.text.x = ggplot2::element_text(size = 10)) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, size = 8))+
  ggplot2::theme(plot.title = element_blank()) +
  NoLegend()
d4
#ggsave(plot = d4, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/10_12_umap_fcm_cell_label_all.png"), width = 5, height = 5)

d4 <- DimPlot(e7386_obj, reduction = "umap", label = F, group.by = "fcm_cell_type") + 
  ggplot2::theme(strip.text.x = ggplot2::element_text(size = 10)) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, size = 8))+
  ggplot2::theme(plot.title = element_blank()) +
  NoLegend()
d4
#ggsave(plot = d4, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/10_12_umap_fcm_cell_label_all_no_label.png"), width = 5, height = 5)


d4 <- DimPlot(e7386_obj, reduction = "umap", label = T, group.by = "fcm_cell_type", split.by = "treatment") + NoLegend()
d4
#ggsave(plot = d4, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/01_09_umap_fcm_cell_label_trt.png"), width = 12, height = 5)
```


```{r}
#combine all B cells into one
e7386_obj_sub$singleRcelltype <- ifelse(e7386_obj_sub$singleRcelltype %in% c('B cells', 'B cells, pro'), "B cells", e7386_obj_sub$singleRcelltype)
#combine all NK cells into one
e7386_obj_sub$singleRcelltype <- ifelse(e7386_obj_sub$singleRcelltype %in% c('NK cells', 'NKT'), "NK cells", e7386_obj_sub$singleRcelltype)

d4 <- DimPlot(e7386_obj_sub, reduction = "umap", label = T, label.box =F, group.by = 'singleRcelltype', label.size = 3.5)
d4
#ggsave(plot = d4, filename = file.path(PATH, "results/4nqo_e7386_images/singleR_all.png"), width = 9, height = 5)

d5 <- DimPlot(e7386_obj_sub, reduction = "umap", split.by = "treatment", label = T, group.by = 'singleRcelltype',  label.size = 3.5, shuffle = T) 
d5
#ggsave(plot = d5, filename = file.path(PATH, "results/4nqo_e7386_images/singleR_all_w_labels.png"), width = 13, height = 5)

e7386_obj_sub$cell_type <- ifelse(e7386_obj_sub$singleRcelltype %in% c("Epithelial cells", "Endothelial cells", "Stromal cells", "Fibroblasts", "Stem cells"), e7386_obj_sub$singleRcelltype, "Immune")
d6 <- DimPlot(e7386_obj_sub, reduction = "umap",  split.by = "treatment",  label = TRUE, group.by = "cell_type", cols = c("tomato", "pink", "green3", "turquoise3", "black", "peru"), label.size = 4) + NoLegend()
d6
#ggsave(plot = d6, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/01_06_umap_coarse_cell_types.png"), width = 13, height = 5)

d6 <- DimPlot(e7386_obj_sub, group.by = "cell_type", reduction = "umap",  label = T, cols = c("tomato", "pink", "green3", "turquoise3", "black", "peru"), label.size = 5) + NoLegend()
d6
#ggsave(plot = d6, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/01_06_umap_coarse_cell_types_nolabel.png"), width = 6, height = 5)
```

```{r}
r1 <- rel_prop(obj = e7386_obj, cond1 = "treatment", cond2 = "fcm_cell_type")
r1
#ggsave(plot = r1, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/01_09_coarse_cell_type_prop.png"), width = 4, height = 5, dpi = 300)
#ggsave(plot = r1, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/10_12_coarse_cell_type_prop.png"), width = 4, height = 5, dpi = 300)

#control
ftest(obj = e7386_obj, samples = c("control", "4NQO control"), cond2 = "treatment", cond1 = "fcm_cell_type")
ftest(obj = e7386_obj, samples = c("control", "4NQO + E7386; 25 mg/kg"), cond2 = "treatment", cond1 = "fcm_cell_type")
ftest(obj = e7386_obj, samples = c("control", "4NQO+ E7386; 50 mg/kg"), cond2 = "treatment", cond1 = "fcm_cell_type")


#4nqo control
ftest(obj = e7386_obj, samples = c("4NQO control", "4NQO + E7386; 25 mg/kg"), cond2 = "treatment", cond1 = "fcm_cell_type")
ftest(obj = e7386_obj, samples = c("4NQO control", "4NQO+ E7386; 50 mg/kg"), cond2 = "treatment", cond1 = "fcm_cell_type")


#e7386
ftest(obj = e7386_obj, samples = c("4NQO + E7386; 25 mg/kg", "4NQO+ E7386; 50 mg/kg"), cond2 = "treatment", cond1 = "fcm_cell_type")

#Fisher's test 
ftest <- fisher.test(table(e7386_obj_sub$cell_type, e7386_obj_sub$treatment), simulate.p.value = T)
ftest
```

```{r}
dat_tab1 <- e7386_obj@meta.data[, c('treatment', 'fcm_cell_type')]
tbl_summary_2 <- tbl_summary(dat_tab1, by = fcm_cell_type, percent = "row")

gt2 <- tbl_summary_2  %>%  as_gt()   # convert to gt table
gt2

# gt2 %>% tab_options(column_labels.font.size=10, table.font.size = 8) %>% gt::gtsave(# save table as image
#     filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/trt_cellTypes.png"),
#   )

#webshot("../mmkhan_e7386/results/e7386_images/trt_cellTypes.png", "mmkhan_e7386/results/e7386_images/trt_cellTypes_adj.png", vwidth  = 1300, vheight = 500)
```

```{r}
table(e7386_obj$singleRcelltype, e7386_obj$fcm_cell_type)
```

## Diff. Exp. Analysis

### Diffanal between clusters
all together cell-type agnostic

```{r eval=FALSE}
Idents(e7386_obj_sub) <- e7386_obj_sub$seurat_clusters
e7386_diffex_clust <- FindAllMarkers(e7386_obj_sub, test.use ="MAST")
e7386_diffex_clust$gene <-  stringr::str_to_title(e7386_diffex_clust$gene)
#saveRDS(e7386_diffex_clust, file.path(PATH, "mmkhan_e7386/results/e7386_diffex_clust_mast.RDS"))
```

```{r}
e7386_diffex_clust <- readRDS(file.path(PATH, "mmkhan_e7386/results/e7386_diffex_clust_mast.RDS"))
diff_genes <- e7386_diffex_clust %>% 
  group_by(cluster) %>%
  filter(avg_log2FC > 1 & p_val_adj <= 0.05) %>%
  dplyr::select(gene) 

diff_genes

diff_list <- as.list(unstack(rev(diff_genes)))
names(diff_list) <- paste("clust", names(diff_list))

diff_list$`clust 2` <- NULL #no DEG set it to null

library(hypeR)

HALLMARK <-  msigdb_gsets("Mus musculus", "H", "")
names(HALLMARK$genesets) <- names(HALLMARK$genesets) %>%  strsplit( "HALLMARK_" ) %>% sapply( tail, 1 )
lmultihyp1 <- hypeR(diff_list, HALLMARK, background = nrow(e7386_obj_sub@assays$RNA@counts))
h1 <- hyp_dots(hyp_obj = lmultihyp1, fdr=0.05, top = 50, merge = TRUE, title = "Hallmark Enrichment", )
h1
#ggsave(plot = h1, filename = file.path(PATH, "results/ctrl_4nqo_ctrl_images/hyper_all.png"), width = 9, height = 5)

REACTOME <- msigdb_gsets(species="Mus musculus", category="C2", subcategory="CP:REACTOME")
names(REACTOME$genesets) <- names(REACTOME$genesets) %>% strsplit( "REACTOME_" ) %>% sapply( tail, 1 )
lmultihyp2 <- hypeR(diff_list, REACTOME, background = nrow(e7386_obj_sub@assays$RNA@counts))
h2 <- hyp_dots(hyp_obj = lmultihyp2, fdr=0.05, top = 30, merge = TRUE, title = "Reactome Enrichment", )
h2

GO <- msigdb_gsets(species="Mus musculus", category="C5", subcategory="BP")
names(GO$genesets) <- names(GO$genesets) %>%  strsplit( "GOBP_" ) %>% sapply( tail, 1 )
lmultihyp3 <- hypeR(diff_list, GO, background = nrow(e7386_obj_sub@assays$RNA@counts))
h3 <- hyp_dots(lmultihyp3, fdr=0.05, merge = TRUE, top = 30, title = "GO Biological Process")
h3

GOMF <- msigdb_gsets(species="Mus musculus", category="C5", subcategory="MF")
names(GOMF$genesets) <- names(GOMF$genesets) %>%  strsplit( "GOMF_" ) %>% sapply( tail, 1 )
lmultihyp4 <- hypeR(diff_list, GOMF, background = nrow(e7386_obj_sub@assays$RNA@counts))
h4 <- hyp_dots(lmultihyp4, fdr=0.05, merge = TRUE, top = 30, title = "GO MF")
h4

GOCC <- msigdb_gsets(species="Mus musculus", category="C5", subcategory="CC")
names(GOCC$genesets) <- names(GOCC$genesets) %>%  strsplit( "GOCC_" ) %>% sapply( tail, 1 )
lmultihyp5 <- hypeR(diff_list, GOCC, background = nrow(e7386_obj_sub@assays$RNA@counts))
h5 <- hyp_dots(lmultihyp5, fdr=0.05, merge = TRUE, top = 30, title = "GO CC")
h5

KEGG <- msigdb_gsets(species="Mus musculus", category = 'C2', subcategory = 'CP:KEGG')
names(KEGG$genesets) <- names(KEGG$genesets) %>%  strsplit( "KEGG_" ) %>% sapply( tail, 1 )
lmultihyp6 <- hypeR(diff_list, KEGG, background = nrow(e7386_obj_sub@assays$RNA@counts))
h6 <- hyp_dots(lmultihyp6, fdr=0.05, merge = TRUE, top = 30, title = "KEGG")
h6

#ggsave(plot = h1, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/hyper_hall_clust.png"), width = 6, height = 6)
#ggsave(plot = h2, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/hyper_react_clust.png"), width = 8, height = 6)
#ggsave(plot = h3, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/hyper_gobp_clust.png"), width = 8, height = 6)
#ggsave(plot = h6, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/hyper_kegg_clust.png"), width = 8, height = 6)


```

### Diffanal between groups
all together cell-type agnostic

```{r eval=F}
Idents(e7386_obj) <- e7386_obj$treatment

#control
e7386_diffex_trt1 <- FindMarkers(e7386_obj, ident.1 = "4NQO control", ident.2 = "control", assay = "RNA", test.use = "MAST")

#4nqo control
e7386_diffex_trt2 <- FindMarkers(e7386_obj, ident.1 = "4NQO + E7386; 25 mg/kg", ident.2 = "4NQO control", assay = "RNA", test.use = "MAST")
e7386_diffex_trt3 <- FindMarkers(e7386_obj, ident.1 = "4NQO+ E7386; 50 mg/kg",  ident.2 = "4NQO control",  assay = "RNA", test.use = "MAST")

#e7386 treated
e7386_diffex_trt4 <- FindMarkers(e7386_obj, ident.1 = "4NQO+ E7386; 50 mg/kg", ident.2 = "4NQO + E7386; 25 mg/kg", test.use ="MAST", assay = "RNA")

e7386_diffex_trt1$gene <-  stringr::str_to_title(rownames(e7386_diffex_trt1))
e7386_diffex_trt2$gene <-  stringr::str_to_title(rownames(e7386_diffex_trt2))
e7386_diffex_trt3$gene <-  stringr::str_to_title(rownames(e7386_diffex_trt3))
e7386_diffex_trt4$gene <-  stringr::str_to_title(rownames(e7386_diffex_trt4))

e7386_diffex_trt1$cond <- rep.int(times  = nrow(e7386_diffex_trt1), x = "4nqo.vs.ctrl")

e7386_diffex_trt2$cond <- rep.int(times  = nrow(e7386_diffex_trt2), x = "e7386@25.vs.4nqo")
e7386_diffex_trt3$cond <- rep.int(times  = nrow(e7386_diffex_trt3), x = "e7386@50.vs.4nqo")

e7386_diffex_trt4$cond <- rep.int(times  = nrow(e7386_diffex_trt4), x = "e7386@50.vs.e7386@25")

e7386_diffex_trt <- rbind(e7386_diffex_trt1, e7386_diffex_trt2, e7386_diffex_trt3, e7386_diffex_trt4)

#saveRDS(e7386_diffex_trt, file.path(PATH, "mmkhan_e7386/results/e7386_diffex_trt.RDS"))
saveRDS(e7386_diffex_trt, file.path(PATH, "mmkhan_e7386/results/2023_10_20_e7386_diffex_trt_all.RDS"))
```

Controlling for cell-type compts

```{r eval=F}
Idents(e7386_obj) <- e7386_obj$treatment

#control
e7386_diffex_trt1 <- FindMarkers(e7386_obj, ident.1 = "4NQO control", ident.2 = "control", assay = "RNA", test.use = "MAST", latent.vars = "fcm_cell_type")

#4nqo control
e7386_diffex_trt2 <- FindMarkers(e7386_obj, ident.1 = "4NQO + E7386; 25 mg/kg", ident.2 = "4NQO control", assay = "RNA", test.use = "MAST", latent.vars = "fcm_cell_type")
e7386_diffex_trt3 <- FindMarkers(e7386_obj, ident.1 = "4NQO+ E7386; 50 mg/kg",  ident.2 = "4NQO control",  assay = "RNA", test.use = "MAST", latent.vars = "fcm_cell_type")

#e7386 treated
e7386_diffex_trt4 <- FindMarkers(e7386_obj, ident.1 = "4NQO+ E7386; 50 mg/kg", ident.2 = "4NQO + E7386; 25 mg/kg", test.use ="MAST", assay = "RNA", latent.vars = "fcm_cell_type")

e7386_diffex_trt1$gene <-  stringr::str_to_title(rownames(e7386_diffex_trt1))
e7386_diffex_trt2$gene <-  stringr::str_to_title(rownames(e7386_diffex_trt2))
e7386_diffex_trt3$gene <-  stringr::str_to_title(rownames(e7386_diffex_trt3))
e7386_diffex_trt4$gene <-  stringr::str_to_title(rownames(e7386_diffex_trt4))

e7386_diffex_trt1$cond <- rep.int(times  = nrow(e7386_diffex_trt1), x = "4nqo.vs.ctrl")
# e7386_diffex_trt2$cond <- rep.int(times  = nrow(e7386_diffex_trt2), x = "ctrl.vs.e7386@25")
# e7386_diffex_trt3$cond <- rep.int(times  = nrow(e7386_diffex_trt3), x = "ctrl.vs.e7386@50")

e7386_diffex_trt2$cond <- rep.int(times  = nrow(e7386_diffex_trt2), x = "e7386@25.vs.4nqo")
e7386_diffex_trt3$cond <- rep.int(times  = nrow(e7386_diffex_trt3), x = "e7386@50.vs.4nqo")

e7386_diffex_trt4$cond <- rep.int(times  = nrow(e7386_diffex_trt4), x = "e7386@50.vs.e7386@25")

e7386_diffex_trt <- rbind(e7386_diffex_trt1, e7386_diffex_trt2, e7386_diffex_trt3, e7386_diffex_trt4)

#saveRDS(e7386_diffex_trt, file.path(PATH, "mmkhan_e7386/results/e7386_diffex_trt.RDS"))
#saveRDS(e7386_diffex_trt, file.path(PATH, "mmkhan_e7386/results/2023_10_20_e7386_diffex_trt_ct_ctrled.RDS"))
```

```{r}
HALLMARK <-  msigdb_gsets("Mus musculus", "H", "")
names(HALLMARK$genesets) <- names(HALLMARK$genesets) %>%  strsplit( "HALLMARK_" ) %>% sapply( tail, 1 )
REACTOME <- msigdb_gsets(species="Mus musculus", category="C2", subcategory="CP:REACTOME")
names(REACTOME$genesets) <- names(REACTOME$genesets) %>% strsplit( "REACTOME_" ) %>% sapply( tail, 1 )
GO <- msigdb_gsets(species="Mus musculus", category="C5", subcategory="BP")
names(GO$genesets) <- names(GO$genesets) %>%  strsplit( "GOBP_" ) %>% sapply( tail, 1 )
GOMF <- msigdb_gsets(species="Mus musculus", category="C5", subcategory="MF")
names(GOMF$genesets) <- names(GOMF$genesets) %>%  strsplit( "GOMF_" ) %>% sapply( tail, 1 )
GOCC <- msigdb_gsets(species="Mus musculus", category="C5", subcategory="CC")
names(GOCC$genesets) <- names(GOCC$genesets) %>%  strsplit( "GOCC_" ) %>% sapply( tail, 1 )
KEGG <- msigdb_gsets(species="Mus musculus", category = 'C2', subcategory = 'CP:KEGG')
names(KEGG$genesets) <- names(KEGG$genesets) %>%  strsplit( "KEGG_" ) %>% sapply( tail, 1 )

```

Pathways with DEGs (not ctrled for CT)

```{r}
e7386_diffex_trt <- readRDS(file.path(PATH, "mmkhan_e7386/results/2023_10_20_e7386_diffex_trt_all.RDS"))
#e7386_diffex_trt <- readRDS(file.path(PATH, "mmkhan_e7386/results/2023_10_20_e7386_diffex_trt_ct_ctrled.RDS"))

e7386_diffex_trt1 <- subset(e7386_diffex_trt, cond=="4nqo.vs.ctrl")
e7386_diffex_trt2 <- subset(e7386_diffex_trt, cond=="e7386@25.vs.4nqo")
e7386_diffex_trt3 <- subset(e7386_diffex_trt, cond=="e7386@50.vs.4nqo")
e7386_diffex_trt4 <- subset(e7386_diffex_trt, cond=="e7386@50.vs.e7386@25")

logfc <- 0.5
fdr <- 0.05

diff_genes1_up <- e7386_diffex_trt1 %>%
  filter(avg_log2FC > logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 

diff_genes1_dn <- e7386_diffex_trt1 %>%
  filter(avg_log2FC < -logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 

diff_genes2_up <- e7386_diffex_trt2 %>%
  filter(avg_log2FC > logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 

diff_genes2_dn <- e7386_diffex_trt2 %>%
 filter(avg_log2FC < -logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 

diff_genes3_up <- e7386_diffex_trt3 %>%
  filter(avg_log2FC > logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 

diff_genes3_dn <- e7386_diffex_trt3 %>%
  filter(avg_log2FC < -logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 

diff_genes4_up <- e7386_diffex_trt4 %>%
  filter(avg_log2FC > logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 

diff_genes4_dn <- e7386_diffex_trt4 %>%
  filter(avg_log2FC < -logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 


diff_list <- list("4nqo.vs.ctrl.up" = diff_genes1_up$gene,
                  "4nqo.vs.ctrl.dn" = diff_genes1_dn$gene,
                  "e7386@25.vs.4nqo.up"=diff_genes2_up$gene,
                  "e7386@25.vs.4nqo.dn"=diff_genes2_dn$gene,
                  "e7386@50.vs.4nqo.up"=diff_genes3_up$gene,
                  "e7386@50.vs.4nqo.dn"=diff_genes3_dn$gene,
                  "e7386@50.vs.e7386@25.up"=diff_genes4_up$gene,
                  "e7386@50.vs.e7386@25.dn"=diff_genes4_dn$gene)


lmultihyp1 <- hypeR(diff_list, HALLMARK, background = nrow(e7386_obj@assays$RNA@counts))
h1 <- hyp_dots(hyp_obj = lmultihyp1, fdr=0.05, top = 20, merge = TRUE, title = "Hallmark Enrichment", )
h1

lmultihyp2 <- hypeR(diff_list, REACTOME, background = nrow(e7386_obj@assays$RNA@counts))
h2 <- hyp_dots(hyp_obj = lmultihyp2, fdr=0.05, top = 50, merge = TRUE, title = "Reactome Enrichment", )
h2


lmultihyp3 <- hypeR(diff_list, GO, background = nrow(e7386_obj@assays$RNA@counts))
h3 <- hyp_dots(lmultihyp3, fdr=0.05, merge = TRUE, top = 20, title = "GO Biological Process")
h3

lmultihyp4 <- hypeR(diff_list, GOMF, background = nrow(e7386_obj@assays$RNA@counts))
h4 <- hyp_dots(lmultihyp4, fdr=0.05, merge = TRUE, top = 50, title = "GO MF", abrv = 75)
h4


lmultihyp5 <- hypeR(diff_list, GOCC, background = nrow(e7386_obj@assays$RNA@counts))
h5 <- hyp_dots(lmultihyp5, fdr=0.05, merge = TRUE, top = 50, title = "GO CC")
h5

lmultihyp6 <- hypeR(diff_list, KEGG, background = nrow(e7386_obj@assays$RNA@counts))
h6 <- hyp_dots(lmultihyp6, fdr=0.05, merge = TRUE, top = 100, title = "KEGG")
h6

# all_gs <- c(unique(HALLMARK$genesets, REACTOME$genesets, GO$genesets, GOMF$genesets, GOCC$genesets, KEGG$genesets))
# lmultihyp7 <- hypeR(diff_list, genesets = all_gs, background = nrow(e7386_obj@assays$RNA@counts))
# h7 <- hyp_dots(lmultihyp7, fdr=0.05, merge = TRUE, top = 50, title = "All pathways")
# h7

#ggsave(plot = h1, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/hyper_hall_trts.png"), width = 6, height = 6)
#ggsave(plot = h2, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/hyper_react_trts.png"), width = 8, height = 6)
#ggsave(plot = h3, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/hyper_gobp_trts.png"), width = 8, height = 6)
```
```{r}
e7386_diffex_trt <- readRDS(file.path(PATH, "mmkhan_e7386/results/2023_10_20_e7386_diffex_trt_ct_ctrled.RDS"))

e7386_diffex_trt1 <- subset(e7386_diffex_trt, cond=="4nqo.vs.ctrl")
e7386_diffex_trt2 <- subset(e7386_diffex_trt, cond=="e7386@25.vs.4nqo")
e7386_diffex_trt3 <- subset(e7386_diffex_trt, cond=="e7386@50.vs.4nqo")
e7386_diffex_trt4 <- subset(e7386_diffex_trt, cond=="e7386@50.vs.e7386@25")

logfc <- 0.5
fdr <- 0.05

diff_genes1_up <- e7386_diffex_trt1 %>%
  filter(avg_log2FC > logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 

diff_genes1_dn <- e7386_diffex_trt1 %>%
  filter(avg_log2FC < -logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 

diff_genes2_up <- e7386_diffex_trt2 %>%
  filter(avg_log2FC > logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 

diff_genes2_dn <- e7386_diffex_trt2 %>%
 filter(avg_log2FC < -logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 

diff_genes3_up <- e7386_diffex_trt3 %>%
  filter(avg_log2FC > logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 

diff_genes3_dn <- e7386_diffex_trt3 %>%
  filter(avg_log2FC < -logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 

diff_genes4_up <- e7386_diffex_trt4 %>%
  filter(avg_log2FC > logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 

diff_genes4_dn <- e7386_diffex_trt4 %>%
  filter(avg_log2FC < -logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 


diff_list <- list("4nqo.vs.ctrl.up" = diff_genes1_up$gene,
                  "4nqo.vs.ctrl.dn" = diff_genes1_dn$gene,
                  "e7386@25.vs.4nqo.up"=diff_genes2_up$gene,
                  "e7386@25.vs.4nqo.dn"=diff_genes2_dn$gene,
                  "e7386@50.vs.4nqo.up"=diff_genes3_up$gene,
                  "e7386@50.vs.4nqo.dn"=diff_genes3_dn$gene,
                  "e7386@50.vs.e7386@25.up"=diff_genes4_up$gene,
                  "e7386@50.vs.e7386@25.dn"=diff_genes4_dn$gene)


lmultihyp1 <- hypeR(diff_list, HALLMARK, background = nrow(e7386_obj@assays$RNA@counts))
h1 <- hyp_dots(hyp_obj = lmultihyp1, fdr=0.05, top = 20, merge = TRUE, title = "Hallmark Enrichment", )
h1

lmultihyp2 <- hypeR(diff_list, REACTOME, background = nrow(e7386_obj@assays$RNA@counts))
h2 <- hyp_dots(hyp_obj = lmultihyp2, fdr=0.05, top = 50, merge = TRUE, title = "Reactome Enrichment", )
h2


lmultihyp3 <- hypeR(diff_list, GO, background = nrow(e7386_obj@assays$RNA@counts))
h3 <- hyp_dots(lmultihyp3, fdr=0.05, merge = TRUE, top = 20, title = "GO Biological Process")
h3

lmultihyp4 <- hypeR(diff_list, GOMF, background = nrow(e7386_obj@assays$RNA@counts))
h4 <- hyp_dots(lmultihyp4, fdr=0.05, merge = TRUE, top = 50, title = "GO MF", abrv = 75)
h4


lmultihyp5 <- hypeR(diff_list, GOCC, background = nrow(e7386_obj@assays$RNA@counts))
h5 <- hyp_dots(lmultihyp5, fdr=0.05, merge = TRUE, top = 50, title = "GO CC")
h5

lmultihyp6 <- hypeR(diff_list, KEGG, background = nrow(e7386_obj@assays$RNA@counts))
h6 <- hyp_dots(lmultihyp6, fdr=0.05, merge = TRUE, top = 100, title = "KEGG")
h6

# all_gs <- c(HALLMARK$genesets, REACTOME$genesets, GO$genesets, GOMF$genesets, GOCC$genesets, KEGG$genesets)
# all_gs_unique <- all_gs[!duplicated(all_gs)]
# lmultihyp7 <- hypeR(diff_list, genesets = all_gs_unique, background = nrow(e7386_obj@assays$RNA@counts))
# h7 <- hyp_dots(lmultihyp7, fdr=0.05, merge = TRUE, abrv = 150, top = 20, title = "All pathways")
# h7

#ggsave(plot = h1, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/hyper_hall_trts.png"), width = 6, height = 6)
#ggsave(plot = h2, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/hyper_react_trts.png"), width = 8, height = 6)
#ggsave(plot = h3, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/hyper_gobp_trts.png"), width = 8, height = 6)
```

## DE analysis 
Random effect: Sample/Mouse

### MAST-Seurat
Might not work as intended for random effects.

```{r eval=F}
Idents(e7386_obj) <- e7386_obj$treatment

#control
e7386_diffex_trt1 <- FindMarkers(e7386_obj, ident.1 = "4NQO control", ident.2 = "control", assay = "RNA", test.use = "MAST", latent.vars = "sample")

#4nqo control
e7386_diffex_trt2 <- FindMarkers(e7386_obj, ident.1 = "4NQO + E7386; 25 mg/kg", ident.2 = "4NQO control", assay = "RNA", test.use = "MAST", latent.vars = "sample")
e7386_diffex_trt3 <- FindMarkers(e7386_obj, ident.1 = "4NQO+ E7386; 50 mg/kg",  ident.2 = "4NQO control",  assay = "RNA", test.use = "MAST", latent.vars = "sample")

#e7386 treated
e7386_diffex_trt4 <- FindMarkers(e7386_obj, ident.1 = "4NQO+ E7386; 50 mg/kg", ident.2 = "4NQO + E7386; 25 mg/kg", test.use ="MAST", assay = "RNA", latent.vars = "sample")

e7386_diffex_trt1$gene <-  stringr::str_to_title(rownames(e7386_diffex_trt1))
e7386_diffex_trt2$gene <-  stringr::str_to_title(rownames(e7386_diffex_trt2))
e7386_diffex_trt3$gene <-  stringr::str_to_title(rownames(e7386_diffex_trt3))
e7386_diffex_trt4$gene <-  stringr::str_to_title(rownames(e7386_diffex_trt4))

e7386_diffex_trt1$cond <- rep.int(times  = nrow(e7386_diffex_trt1), x = "4nqo.vs.ctrl")
# e7386_diffex_trt2$cond <- rep.int(times  = nrow(e7386_diffex_trt2), x = "ctrl.vs.e7386@25")
# e7386_diffex_trt3$cond <- rep.int(times  = nrow(e7386_diffex_trt3), x = "ctrl.vs.e7386@50")

e7386_diffex_trt2$cond <- rep.int(times  = nrow(e7386_diffex_trt2), x = "e7386@25.vs.4nqo")
e7386_diffex_trt3$cond <- rep.int(times  = nrow(e7386_diffex_trt3), x = "e7386@50.vs.4nqo")

e7386_diffex_trt4$cond <- rep.int(times  = nrow(e7386_diffex_trt4), x = "e7386@50.vs.e7386@25")

e7386_diffex_trt <- rbind(e7386_diffex_trt1, e7386_diffex_trt2, e7386_diffex_trt3, e7386_diffex_trt4)

#saveRDS(e7386_diffex_trt, file.path(PATH, "mmkhan_e7386/results/e7386_diffex_trt.RDS"))
#saveRDS(e7386_diffex_trt, file.path(PATH, "mmkhan_e7386/results/2023_10_20_e7386_diffex_trt_ct_ctrled.RDS"))
```

```{r}
HALLMARK <-  msigdb_gsets("Mus musculus", "H", "")
names(HALLMARK$genesets) <- names(HALLMARK$genesets) %>%  strsplit( "HALLMARK_" ) %>% sapply( tail, 1 )
REACTOME <- msigdb_gsets(species="Mus musculus", category="C2", subcategory="CP:REACTOME")
names(REACTOME$genesets) <- names(REACTOME$genesets) %>% strsplit( "REACTOME_" ) %>% sapply( tail, 1 )
GO <- msigdb_gsets(species="Mus musculus", category="C5", subcategory="BP")
names(GO$genesets) <- names(GO$genesets) %>%  strsplit( "GOBP_" ) %>% sapply( tail, 1 )
GOMF <- msigdb_gsets(species="Mus musculus", category="C5", subcategory="MF")
names(GOMF$genesets) <- names(GOMF$genesets) %>%  strsplit( "GOMF_" ) %>% sapply( tail, 1 )
GOCC <- msigdb_gsets(species="Mus musculus", category="C5", subcategory="CC")
names(GOCC$genesets) <- names(GOCC$genesets) %>%  strsplit( "GOCC_" ) %>% sapply( tail, 1 )
KEGG <- msigdb_gsets(species="Mus musculus", category = 'C2', subcategory = 'CP:KEGG')
names(KEGG$genesets) <- names(KEGG$genesets) %>%  strsplit( "KEGG_" ) %>% sapply( tail, 1 )

```

Pathways with DEGs

```{r fig.width=15, fig.height=12}
e7386_diffex_trt <- readRDS(file.path(PATH, "mmkhan_e7386/results/2023_10_20_e7386_diffex_trt_all.RDS"))
#e7386_diffex_trt <- readRDS(file.path(PATH, "mmkhan_e7386/results/2023_10_20_e7386_diffex_trt_ct_ctrled.RDS"))

e7386_diffex_trt1 <- subset(e7386_diffex_trt, cond=="4nqo.vs.ctrl")
e7386_diffex_trt2 <- subset(e7386_diffex_trt, cond=="e7386@25.vs.4nqo")
e7386_diffex_trt3 <- subset(e7386_diffex_trt, cond=="e7386@50.vs.4nqo")
e7386_diffex_trt4 <- subset(e7386_diffex_trt, cond=="e7386@50.vs.e7386@25")

logfc <- 0.5
fdr <- 0.05

diff_genes1_up <- e7386_diffex_trt1 %>%
  filter(avg_log2FC > logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 

diff_genes1_dn <- e7386_diffex_trt1 %>%
  filter(avg_log2FC < -logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 

diff_genes2_up <- e7386_diffex_trt2 %>%
  filter(avg_log2FC > logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 

diff_genes2_dn <- e7386_diffex_trt2 %>%
 filter(avg_log2FC < -logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 

diff_genes3_up <- e7386_diffex_trt3 %>%
  filter(avg_log2FC > logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 

diff_genes3_dn <- e7386_diffex_trt3 %>%
  filter(avg_log2FC < -logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 

diff_genes4_up <- e7386_diffex_trt4 %>%
  filter(avg_log2FC > logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 

diff_genes4_dn <- e7386_diffex_trt4 %>%
  filter(avg_log2FC < -logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 


diff_list <- list("4nqo.vs.ctrl.up" = diff_genes1_up$gene,
                  "4nqo.vs.ctrl.dn" = diff_genes1_dn$gene,
                  "e7386@25.vs.4nqo.up"=diff_genes2_up$gene,
                  "e7386@25.vs.4nqo.dn"=diff_genes2_dn$gene,
                  "e7386@50.vs.4nqo.up"=diff_genes3_up$gene,
                  "e7386@50.vs.4nqo.dn"=diff_genes3_dn$gene,
                  "e7386@50.vs.e7386@25.up"=diff_genes4_up$gene,
                  "e7386@50.vs.e7386@25.dn"=diff_genes4_dn$gene)


lmultihyp1 <- hypeR(diff_list, HALLMARK, background = nrow(e7386_obj@assays$RNA@counts))
h1 <- hyp_dots(hyp_obj = lmultihyp1, fdr=0.05, top = 20, merge = TRUE, title = "Hallmark Enrichment", )
h1

lmultihyp2 <- hypeR(diff_list, REACTOME, background = nrow(e7386_obj@assays$RNA@counts))
h2 <- hyp_dots(hyp_obj = lmultihyp2, fdr=0.05, top = 50, merge = TRUE, title = "Reactome Enrichment", )
h2


lmultihyp3 <- hypeR(diff_list, GO, background = nrow(e7386_obj@assays$RNA@counts))
h3 <- hyp_dots(lmultihyp3, fdr=0.05, merge = TRUE, top = 20, title = "GO Biological Process")
h3

lmultihyp4 <- hypeR(diff_list, GOMF, background = nrow(e7386_obj@assays$RNA@counts))
h4 <- hyp_dots(lmultihyp4, fdr=0.05, merge = TRUE, top = 50, title = "GO MF", abrv = 75)
h4


lmultihyp5 <- hypeR(diff_list, GOCC, background = nrow(e7386_obj@assays$RNA@counts))
h5 <- hyp_dots(lmultihyp5, fdr=0.05, merge = TRUE, top = 50, title = "GO CC")
h5

lmultihyp6 <- hypeR(diff_list, KEGG, background = nrow(e7386_obj@assays$RNA@counts))
h6 <- hyp_dots(lmultihyp6, fdr=0.05, merge = TRUE, top = 100, title = "KEGG")
h6

# all_gs <- c(unique(HALLMARK$genesets, REACTOME$genesets, GO$genesets, GOMF$genesets, GOCC$genesets, KEGG$genesets))
# lmultihyp7 <- hypeR(diff_list, genesets = all_gs, background = nrow(e7386_obj@assays$RNA@counts))
# h7 <- hyp_dots(lmultihyp7, fdr=0.05, merge = TRUE, top = 50, title = "All pathways")
# h7

#ggsave(plot = h1, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/hyper_hall_trts.png"), width = 6, height = 6)
#ggsave(plot = h2, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/hyper_react_trts.png"), width = 8, height = 6)
#ggsave(plot = h3, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/hyper_gobp_trts.png"), width = 8, height = 6)
```
### MAST-wrapper

```{r eval=F}
.FoldChange.default <- function(
  data,
  cells.1,
  cells.2,
  mean.fxn,
  fc.name,
  features = NULL,
  ...
) {

  features <- if (is.null(features)) rownames(data) else features
  # Calculate percent expressed
  thresh.min <- 0
  pct.1 <- round(
    x = rowSums(x = data[rownames(data) %in% features, colnames(data) %in% cells.1, drop = FALSE] > thresh.min) /
      length(x = cells.1),
    digits = 3
  )
  pct.2 <- round(
    x = rowSums(x = data[rownames(data) %in% features, colnames(data) %in% cells.2, drop = FALSE] > thresh.min) /
      length(x = cells.2),
    digits = 3
  )
  # Calculate fold change
  data.1 <- mean.fxn(data[rownames(data) %in% features, colnames(data) %in% cells.1, drop = FALSE])
  data.2 <- mean.fxn(data[rownames(data) %in% features, colnames(data) %in% cells.2, drop = FALSE])
  fc <- (data.1 - data.2)
  fc.results <- as.data.frame(x = cbind(fc, pct.1, pct.2))
  colnames(fc.results) <- c(fc.name, "pct.1", "pct.2")
  return(fc.results)
}
#' MAST differential expression test with random effect for Seurat object
#'
#' Uses Seurat::FindMarkers MAST test hurdle model with added random effects variables
#'
#' @param object Seurat >=3 object, with log-normalized data in the `data` slot
#' @param random_effect.vars Character vector of variables to add as random intersects in MAST model i.e. ~ ...  + (1|random_effect.var1) + (1|random_effect.var2)
#' @param ident.1 Identity class to define markers for
#' @param ident.2 A second identity class for comparison. Leave as NULL to compare with all other cells
#' @param cells.1 Vector of cell names belonging to group 1. Alternative way to specify ident.1
#' @param cells.2 Vector of cell names belonging to group 2. Alternative way to specify ident.2
#' @param logfc.threshold Only return results with a DE exceeding threshold
#' @param base base for log when computing mean and for output, default exp(1). NB: Seurat has changed to log2 in V4.
#' @param group.by Regroup cells into a different identity class prior to performing differential expression
#' @param assay Assay to pull data from; defaults to default assay
#' @param slot Slot to pull data from; defaults to "data" as MAST expects log-normalized data
#' @param features Genes to test. Default is to use all genes
#' @param min.pct Only test genes that are detected in a minimum fraction of min.pct cells in either of the two populations. Meant to speed up the function by not testing genes that are very infrequently expressed. Default is 0.1
#' @param max.cells.per.ident Downsample each identity class to a max number. Default is no downsampling. Not activated by default (set to Inf)
#' @param random.seed Random seed to use for down sampling
#' @param latent.vars Variables to test
#' @param n_cores How many cores to use (temporarily sets options(mc.cores=n_cores))
#' @param verbose whether to print stdout from MAST functions
#' @param p.adjust.method passed to p.adjust. Note that n is all the genes in the object
#' @param \dots Additional parameters (other than formula, sca, method, ebayes, strictConvergence) to pass to MAST::zlm
#' @return data.frame with column names p_val, avg_log[base]FC, pct.1, pct.2, p_val_adj (identical format to Seurat::FindMarkers)
#' @export
#'
#' @references Zimmerman, K.D., Espeland, M.A. & Langefeld, C.D.
#' . A practical solution to pseudoreplication bias in single-cell studies.
#' . Nat Commun 12, 738 (2021). https://doi.org/10.1038/s41467-021-21038-1
#'
#' . McDavid A, Finak G, Yajima M (2020). MAST: Model-based Analysis of Single Cell Transcriptomics. R package version 1.16.0, https://github.com/RGLab/MAST/.
#'
#' . Stuart and Butler et al. Comprehensive Integration of Single-Cell Data. Cell (2019)
#'
#' . Seurat 3 source at https://github.com/satijalab/seurat/blob/master/R/differential_expression.R
DE_MAST_RE_seurat = function(
  object,
  random_effect.vars,
  ident.1 = NULL,
  ident.2 = NULL,
  cells.1 = NULL,
  cells.2 = NULL,
  group.by = NULL,
  logfc.threshold = 0.25,
  base = exp(1),
  assay=NULL,
  slot="data",
  features = NULL,
  min.pct = 0.1,
  max.cells.per.ident = NULL,
  random.seed = 1,
  latent.vars = NULL,
  n_cores=NULL,
  verbose=TRUE,
  p.adjust.method="fdr",
  ...
) {

  # require(Seurat)
  # require(MAST)

  if (!is.null(n_cores)) {
    # https://www.tidyverse.org/blog/2020/04/self-cleaning-test-fixtures/#the-onexit-pattern
    # set new value and capture old in op
    op <- options("mc.cores"=n_cores)
    on.exit(options(op), add = T)
  }

  if (is.null(c(ident.1, ident.2, cells.1, cells.2))) stop("at least one of ident.1, ident.2, cells.1, cells.2 must be provided")
  if (!is.null(ident.1) & !is.null(cells.1)) stop("Provide one of ident.1 or cells.1 but not both")
  if (!is.null(ident.2) & !is.null(cells.2)) stop("Provide one of ident.2 or cells.2 but not both")

  fc.name  <- if (base == exp(1)) "avg_logFC" else paste0("avg_log", base, "FC")
  #======== check inputs ========================================

  stopifnot(!is.null(random_effect.vars))
  stopifnot(all(random_effect.vars %in% colnames(object@meta.data)))

  if (slot != "data") warning(paste0("MAST uses the logNormalised counts which are usually in the 'data' slot, but you are using ",slot))

  if (verbose) message(paste0("using ", round(base,2), " as log base. Make sure that the data slot has been log-transformed using this base"))
  #======== resolve idents and get cells ===================

  anyNA = if (!is.null(group.by)) any(is.na(object@meta.data[[group.by]])) else any(is.na(Seurat::Idents(object)))
  if (anyNA) stop("Some identities are NA, please check the metadata")

  logical.cells.1 = if (!is.null(cells.1)) colnames(object) %in% cells.1 else { if (!is.null(group.by)) {object@meta.data[[group.by]] == ident.1} else {Seurat::Idents(object) == ident.1}}
  if (sum(logical.cells.1)==0) {
    stop("no cells found matching ident.1. Did you forget to set Idents(object) or use group.by?")
  }
  if (is.null(cells.1)) cells.1 = colnames(object)[logical.cells.1]

  logical.cells.2 = if (!is.null(cells.2)) colnames(object) %in% cells.2 else {if (is.null(ident.2)) {!logical.cells.1} else {if (!is.null(group.by)) {object@meta.data[[group.by]] == ident.2} else {Seurat::Idents(object) == ident.2}}}
  if (sum(logical.cells.2)==0) {
    stop("no cells found matching ident.2. Did you forget to set Idents(object) or use group.by?")
  }
  if (is.null(cells.2)) cells.2 = colnames(object)[logical.cells.2]

  #======== features =================================

  features <- if (is.null(features)) rownames(object) else features

  if (any(!features %in% rownames(object))) {
    warning(paste0(paste(features[!features %in% rownames(object)], collapse = ", "), " not found in data!"))
  }

  features = features[features %in% rownames(object)]
  vec_logical_features = rep(T, length(features))

  data = Seurat::GetAssayData(object = object, assay=assay, slot=slot)

  densemat = utils_big_as.matrix(data, n_slices_init = 1, verbose=F)

  # compute average log fold change
  pseudocount.use = 1
  mean.fxn = function(x) {
    return(log(x = rowMeans(x = expm1(x = x)) + pseudocount.use, base = base))
  }
  # outputs a data.frame with columns fc.name, "pct.1", "pct.2"
  fc.results = .FoldChange.default(
      data=densemat,
      cells.1=cells.1,
      cells.2=cells.2,
      mean.fxn=mean.fxn,
      fc.name=fc.name,
      features = features)

  if (!is.null(logfc.threshold)) {
    vec_logical_features = vec_logical_features & fc.results[[fc.name]]>logfc.threshold  # need to convert back to non-log space to take mean
  }
  if (sum(vec_logical_features)<2) stop(paste0("Fewer than two features have a log fold change above ", logfc.threshold))
  if (!is.null(min.pct)) {
    vec_logical_features = vec_logical_features & (fc.results$pct.1 > min.pct | fc.results$pct.2 > min.pct)
  }
  if (sum(vec_logical_features)<2) stop(paste0("Fewer than two features are expressed in ", min.pct, " of cells in any condition"))

  features <- features[vec_logical_features]

  densemat = densemat[rownames(densemat) %in% features, , drop=F]

  #======== down sample cells ==============================

  idx.cells.1 = which(logical.cells.1)
  idx.cells.2 = which(logical.cells.2)

  if (!is.null(max.cells.per.ident)) {
    if (sum(logical.cells.1) > max.cells.per.ident) {
      set.seed(randomSeed)
      idx.cells.1 = sample(x = idx.cells.1, size = max.cells.per.ident, replace = F)
    }
    if (sum(!logical.cells.1) > max.cells.per.ident) {
      set.seed(randomSeed)
      idx.cells.2 = sample(x = idx.cells.2, size = max.cells.per.ident, replace = F)
    }
  }

  cells.1 = colnames(densemat)[idx.cells.1]
  cells.2 = colnames(densemat)[idx.cells.2]
  densemat = densemat[,c(idx.cells.1,idx.cells.2), drop=F]

  #======== filter out all zero features (if not already removed) ==

  densemat = densemat[apply(X = densemat, MARGIN=1, FUN = sum)>0,]

  #======== prep column (cell/sample) data =================

  df_coldata = data.frame(
    row.names = c(cells.1, cells.2),
    "wellKey" = c(cells.1, cells.2)
    ) # wellKey is hardcoded feature name in MAST

  # DE group factor
  df_coldata[cells.1, "group"] <- "Group1"
  df_coldata[cells.2, "group"] <- "Group2"
  df_coldata[, "group"] <- factor(x = df_coldata[, "group"])

  # latent vars
  if (!is.null(latent.vars)) {
    for (latent.var in latent.vars) {
      df_coldata[[latent.var]] <- object@meta.data[c(idx.cells.1,idx.cells.2), latent.var]
    }
  }

  # add random vars to column data
  for (random_effect.var in random_effect.vars) {
    df_coldata[[random_effect.var]] <- as.factor(
        object@meta.data[c(idx.cells.1,idx.cells.2),random_effect.var]
      )
  }

  # check for NAs in column data
  for (colname in colnames(df_coldata)) {
    if (any(is.na(df_coldata[[colname]]))) {
      stop(paste0("variable '", colname, "' contains NAs"))
    }
  }

  # check whether random var levels overlap entirely with test condition
  for (random_effect.var in random_effect.vars) {
    for (re_lvl in levels(df_coldata[[random_effect.var]])) {
      if (all(df_coldata[[random_effect.var]]==re_lvl & 1:ncol(densemat) %in% idx.cells.1)) {
        stop(paste0(random_effect.var, " level ", re_lvl, " overlaps entirely with ident.1. Increase ", max.cells.per.ident, " or check the data"))
      }
      else if (all(df_coldata[[random_effect.var]]==re_lvl & 1:ncol(densemat) %in% idx.cells.2)) {
        stop(paste0(random_effect.var, " level ", re_lvl, " overlaps entirely with ident.2. Increase ", max.cells.per.ident, " or check the data"))
      }
    }
  }

  #======== prep  feature data =========================

  df_feature = data.frame("primerid"=rownames(densemat)) # primerid is hardcoded feature name in MAST

  #======== prep SingleCellAssay object ============================

  expr = quote(MAST::FromMatrix(exprsArray=densemat,
                                 check_sanity = TRUE,
                                 cData=df_coldata,
                                 fData=df_feature))
  sca <- if (verbose) eval(expr) else suppressMessages(eval(expr))

  # set group as a factor and make group1 the reference
  #cond <- factor(x = SummarizedExperiment::colData(sca)$group)
  # in Seurat source this is Group1!?
  # see https://github.com/satijalab/seurat/blob/master/R/differential_expression.R
  SummarizedExperiment::colData(sca)$group <- relevel(SummarizedExperiment::colData(sca)$group, ref="Group2")
  # cond <- relevel(x = cond, ref = "Group2")
  # SummarizedExperiment::colData(sca)$condition <- cond

  # add the centered number of expressed genes to the model
  # cdr2 <- colSums(SummarizedExperiment::assay(sca)>0)
  # SummarizedExperiment::colData(sca)$cngeneson <- scale(cdr2)

  #======== run MAST test ======================

  str_plus = if (!is.null(latent.vars)) " + " else ""

  fmla <- as.formula(
    object = paste0(" ~ group", str_plus, paste(latent.vars, collapse = " + "), paste(" + (1 |", random_effect.vars, ")", collapse=""))
  )

  # fit model parameters
  expr = quote(MAST::zlm(formula = fmla,
                   sca = sca,
                   method = 'glmer',
                   ebayes = F,
                   strictConvergence = FALSE,
                   ...))

  zlmCond <- if (verbose) eval(expr) else suppressMessages(eval(expr))

  if (verbose) {
    print(zlmCond)
    message("Compute likelihoods and p-values")
  }
  # call a likelihood ratio test on the fitted object
  expr = quote(MAST::summary(object = zlmCond, doLRT = 'groupGroup1'))
  summaryCond <- if (verbose) eval(expr) else suppressWarnings(suppressMessages(eval(expr)))

  if ("character" %in% class(summaryCond)) stop("No differentially genes detected")
  # print(summaryCond,n=4)
  # Fitted zlm with top 4 genes per contrast:
  #   ( log fold change Z-score )
  # primerid           conditionGroup1 cngeneson
  # Akr1c20              -4.1            13.2*
  # Bclaf3               18.5*            4.8
  # Cmss1                -0.3            -9.9*
  # Cux2                -16.5*            4.6
  # ENSMUSG00000115022  -19.5*            7.6
  # Pot1b                13.5*            5.3
  # Siah1a                4.6            17.6*
  # Sipa1l2               1.8            10.5*
  #
  #
  # summaryCond$datatable has colnames
  # "primerid" feature name
  # "component" if no latent.vars, "C" (continuous) "D" (discrete)  "H" (hurdle) "S" (RE?) "logFC"
  # "contrast" if no latent.vars,  conditionGroup1 (Intercept) cngeneson
  # "ci.hi" confidence interval (high)
  # "ci.lo" confidence interval (low)
  # "coef" coefficient estimate
  # "z" z-scores

  # uses
  # https://www.bioconductor.org/packages/release/bioc/vignettes/MAST/inst/doc/MAITAnalysis.html#differential-expression-using-a-hurdle-model
  # and
  # https://github.com/kdzimm/PseudoreplicationPaper/blob/master/Type_1_Error/Type%201%20-%20MAST%20RE.Rmd

  de.results <- data.frame(
    "p_val" = summaryCond$datatable[contrast=='groupGroup1' & component=='H', `Pr(>Chisq)`],
    fc.results[vec_logical_features,])  #setDF(summaryCond$datatable[contrast=='groupGroup1' & component=='logFC', .(coef)])

  de.results$p_val_adj = p.adjust(de.results$p_val, method=p.adjust.method, n=length(vec_logical_features))

  de.results = de.results[order(de.results$p_val, -de.results[[fc.name]]),]

  return(de.results)
}


```

```{r eval=F}
library(SCOPfunctions)
diff1 <- DE_MAST_RE_seurat(object = e7386_obj, random_effect.vars = "sample", ident.1 = "4NQO control", ident.2 = "control", assay = "RNA")

#4nqo control
diff2 <- DE_MAST_RE_seurat(e7386_obj, ident.1 = "4NQO + E7386; 25 mg/kg", ident.2 = "4NQO control", random_effect.vars = "sample", assay = "RNA")
diff3 <- DE_MAST_RE_seurat(e7386_obj, ident.1 = "4NQO+ E7386; 50 mg/kg",  ident.2 = "4NQO control", random_effect.vars = "sample", assay = "RNA")

diff1$gene <-  stringr::str_to_title(rownames(diff1))
diff2$gene <-  stringr::str_to_title(rownames(diff2))
diff3$gene <-  stringr::str_to_title(rownames(diff3))

diff1$cond <- rep.int(times  = nrow(diff1), x = "4nqo.vs.ctrl")
# e7386_diffex_trt2$cond <- rep.int(times  = nrow(e7386_diffex_trt2), x = "ctrl.vs.e7386@25")
# e7386_diffex_trt3$cond <- rep.int(times  = nrow(e7386_diffex_trt3), x = "ctrl.vs.e7386@50")

diff2$cond <- rep.int(times  = nrow(diff2), x = "e7386@25.vs.4nqo")
diff3$cond <- rep.int(times  = nrow(diff3), x = "e7386@50.vs.4nqo")

e7386_diffex_trt <- rbind(diff1, diff2, diff3)
```





```{r fig.width=15, fig.height=12}
e7386_diffex_trt <- readRDS(file.path(PATH, "mmkhan_e7386/results/2023_10_20_e7386_diffex_trt_ct_ctrled.RDS"))

e7386_diffex_trt1 <- subset(e7386_diffex_trt, cond=="4nqo.vs.ctrl")
e7386_diffex_trt2 <- subset(e7386_diffex_trt, cond=="e7386@25.vs.4nqo")
e7386_diffex_trt3 <- subset(e7386_diffex_trt, cond=="e7386@50.vs.4nqo")
e7386_diffex_trt4 <- subset(e7386_diffex_trt, cond=="e7386@50.vs.e7386@25")

logfc <- 0
fdr <- 0.05

diff_genes1_up <- e7386_diffex_trt1 %>%
  filter(avg_log2FC > logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 

diff_genes1_dn <- e7386_diffex_trt1 %>%
  filter(avg_log2FC < -logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 

diff_genes2_up <- e7386_diffex_trt2 %>%
  filter(avg_log2FC > logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 

diff_genes2_dn <- e7386_diffex_trt2 %>%
 filter(avg_log2FC < -logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 

diff_genes3_up <- e7386_diffex_trt3 %>%
  filter(avg_log2FC > logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 

diff_genes3_dn <- e7386_diffex_trt3 %>%
  filter(avg_log2FC < -logfc & p_val_adj <= fdr) %>%
  dplyr::select(gene) 


diff_list <- list("4nqo.vs.ctrl.up" = diff_genes1_up$gene,
                  "4nqo.vs.ctrl.dn" = diff_genes1_dn$gene,
                  "e7386@25.vs.4nqo.up"=diff_genes2_up$gene,
                  "e7386@25.vs.4nqo.dn"=diff_genes2_dn$gene,
                  "e7386@50.vs.4nqo.up"=diff_genes3_up$gene,
                  "e7386@50.vs.4nqo.dn"=diff_genes3_dn$gene
                 )


lmultihyp1 <- hypeR(diff_list, HALLMARK, background = nrow(e7386_obj@assays$RNA@counts))
h1 <- hyp_dots(hyp_obj = lmultihyp1, fdr=0.05, top = 20, merge = TRUE, title = "Hallmark Enrichment", )
h1

lmultihyp2 <- hypeR(diff_list, REACTOME, background = nrow(e7386_obj@assays$RNA@counts))
h2 <- hyp_dots(hyp_obj = lmultihyp2, fdr=0.05, top = 50, merge = TRUE, title = "Reactome Enrichment", )
h2


lmultihyp3 <- hypeR(diff_list, GO, background = nrow(e7386_obj@assays$RNA@counts))
h3 <- hyp_dots(lmultihyp3, fdr=0.05, merge = TRUE, top = 20, title = "GO Biological Process")
h3

lmultihyp4 <- hypeR(diff_list, GOMF, background = nrow(e7386_obj@assays$RNA@counts))
h4 <- hyp_dots(lmultihyp4, fdr=0.05, merge = TRUE, top = 50, title = "GO MF", abrv = 75)
h4


lmultihyp5 <- hypeR(diff_list, GOCC, background = nrow(e7386_obj@assays$RNA@counts))
h5 <- hyp_dots(lmultihyp5, fdr=0.05, merge = TRUE, top = 50, title = "GO CC")
h5

lmultihyp6 <- hypeR(diff_list, KEGG, background = nrow(e7386_obj@assays$RNA@counts))
h6 <- hyp_dots(lmultihyp6, fdr=0.05, merge = TRUE, top = 100, title = "KEGG")
h6

# all_gs <- c(HALLMARK$genesets, REACTOME$genesets, GO$genesets, GOMF$genesets, GOCC$genesets, KEGG$genesets)
# all_gs_unique <- all_gs[!duplicated(all_gs)]
# lmultihyp7 <- hypeR(diff_list, genesets = all_gs_unique, background = nrow(e7386_obj@assays$RNA@counts))
# h7 <- hyp_dots(lmultihyp7, fdr=0.05, merge = TRUE, abrv = 150, top = 20, title = "All pathways")
# h7

#ggsave(plot = h1, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/hyper_hall_trts.png"), width = 6, height = 6)
#ggsave(plot = h2, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/hyper_react_trts.png"), width = 8, height = 6)
#ggsave(plot = h3, filename = file.path(PATH, "mmkhan_e7386/results/e7386_images/hyper_gobp_trts.png"), width = 8, height = 6)
```

## Misc.

### Epithelial

```{r eval=F}
FeaturePlot(object = e7386_obj, features = c("CDH1", "KRT5", "VIM"), split.by = "treatment")
FeaturePlot(object = e7386_obj, features = c("KRT5", "KRT6", "KRT19")) #KRT genes

ge_boxplot(obj = e7386_obj, genes = c("CDH1", "KRT5", "VIM"), cond = "treatment", ncol = 3) 
ge_boxplot(obj = e7386_obj, genes = c("CD70", "PDPN"), cond = "treatment", ncol = 2) 
ge_boxplot(obj = e7386_obj, genes = c("KRT5", "KRT6", "KRT19"), cond = "treatment", ncol = 3) 

ge_boxplot(obj = e7386_obj, genes = c("FAP", "PDGFRA"), cond = "treatment", ncol = 2)
ge_boxplot(genes = c("MCAM", "CAV1"), obj = e7386_obj, cond = "treatment", ncol = 2)


FeaturePlot(object = e7386_obj, features = c("EGFR", "PDGFRB"))

FeaturePlot(object = e7386_obj, features = c("KRT8", "KRT18"), split.by = "treatment")
FeaturePlot(object = e7386_obj, features = c("PDPN"), split.by = "treatment")

FeaturePlot(object = e7386_obj, features = c("PTPRC"), split.by = "treatment")
```

### CAFs
[Biomarkers for cancer-associated fibroblasts](https://biomarkerres.biomedcentral.com/articles/10.1186/s40364-020-00245-w)
### Neutral

```{r eval=F}
FeaturePlot(object = e7386_obj, features = c("ACTA2", "S100A4"), split.by = "treatment")
ge_boxplot(obj = e7386_obj, genes = c("ACTA2", "S100A4"), cond = "treatment", ncol = 2)
```

#### Pro

```{r eval=F}
FeaturePlot(object = e7386_obj, features = c("FAP", "PDGFRA"), split.by = "treatment")
FeaturePlot(object = e7386_obj, features = c("CD70", "PDPN"), split.by = "treatment")
FeaturePlot(object = e7386_obj, features = c("C5AR2", "MME", "CD74"), split.by = "treatment")
```

#### Suppressors

```{r eval=F}
FeaturePlot(object = e7386_obj, features = c("MCAM", "CAV1"), split.by = "treatment")
VlnPlot(object = e7386_obj, features = c("MCAM", "CAV1"), group.by = "treatment")
```
```{r eval=F}
ge_boxplot(obj = e7386_obj, genes = c("CDH1", "KRT5", "VIM"), cond = "treatment", ncol = 3) 
ge_boxplot(obj = e7386_obj, genes = c("CD70", "PDPN"), cond = "treatment", ncol = 2) 
ge_boxplot(obj = e7386_obj, genes = c("FAP", "PDGFRA"), cond = "treatment", ncol = 2)
ge_boxplot(genes = c("MCAM", "CAV1"), obj = e7386_obj, cond = "treatment", ncol = 2)
```

### Stress genes

```{r}
ge_boxplot <- function(genes, obj, cond, ncol=1){
  if(length(genes)==1) {
    df <- as.data.frame(as.matrix(obj@assays$RNA@data[c(genes),]))
    colnames(df) <- c(genes)
  } else {
      df <- as.data.frame(t(as.matrix(obj@assays$RNA@data[c(genes),])))
  }
  condition <- cond
  df[, paste(cond)] <- obj@meta.data[, paste(cond)]
  df[, paste(cond)] <- factor(df[, paste(cond)])

  gsvaResFlat <- reshape2::melt(df, id.vars = paste(cond, collapse = "_"), variable.name = "group")

  g1 <- ggplot2::ggplot(gsvaResFlat, ggplot2::aes_string(
    x = paste(condition, collapse = "_"), y = "value",
    color = paste(condition, collapse = "_"))) +
    ggplot2::geom_boxplot(outlier.shape = NA) +
    #ggplot2::geom_jitter(size=0.0) +
    ggplot2::facet_wrap(~group, scale = "free_y",
                        ncol = ncol) +
    ggpubr::stat_compare_means(method = "anova",  label = 'p.format')+
    ggplot2::theme(strip.text.x = ggplot2::element_text(size = 10))+
    ggplot2::theme(axis.text.x = element_text(angle = 45, hjust = 1))+
    labs(title = 'Gene Expression', y = "log-Norm Counts", x = cond)+
    theme(axis.text = element_text(size = 10), axis.title = element_text(size = 10, face = "bold"))
  g1
}
```

```{r}
ge_boxplot(genes = c('JUN', 'FOS', 'JUNB', 'FOSB'), obj = e7386_obj, cond = "fcm_cell_type", ncol = 2)
ge_boxplot(genes = c('JUN', 'FOS', 'JUNB', 'FOSB'), obj = e7386_obj, cond = "fcm_cell_label", ncol = 2)
```

### PML sigs

```{r}
gsva_res_all <- data.frame(t(readRDS(file.path(PATH, "mmkhan_e7386/results/2023_07_21_e7386_all_pml_sigs.rds"))))
gsva_res_all <- gsva_res_all[,c('hknr_up', 'dys_up', 'cancer_up', 'hknr_dn', 'dys_dn', 'cancer_dn')]

gsva_res_all$cancer_diff <- gsva_res_all$cancer_up - gsva_res_all$cancer_dn
gsva_res_all$dys_diff <- gsva_res_all$dys_up - gsva_res_all$dys_dn
gsva_res_all$hknr_diff <- gsva_res_all$hknr_up - gsva_res_all$hknr_dn
```

```{r}
gsva_res_all$treatment <- e7386_obj$treatment[match(rownames(gsva_res_all), e7386_obj$cell_id)]
gsva_res_all$seurat_clusters <- e7386_obj$seurat_clusters[match(rownames(gsva_res_all), e7386_obj$cell_id)]
gsva_res_all$fcm_cell_type <- e7386_obj$fcm_cell_type[match(rownames(gsva_res_all), e7386_obj$cell_id)]
gsva_res_all$fcm_cell_label <- e7386_obj$fcm_cell_label[match(rownames(gsva_res_all), e7386_obj$cell_id)]
```

#### Diff scores

```{r}
gsva_res_all_diff <- gsva_res_all[,c('hknr_diff', 'dys_diff', 'cancer_diff', 'treatment', 'seurat_clusters', 'fcm_cell_type', 'fcm_cell_label')]
```

```{r}
gsva_boxplot <- function(df, group, nc, cls) {
print(group)
gsvaResT <- df[, colnames(df) %in% c(cls, paste(group))]
condition <- group 
cond <- as.list(gsvaResT[, condition ,drop = FALSE])
gsvaResT[, paste(condition, collapse = "_")] <- cond
gsvaResFlat <- reshape2::melt(gsvaResT, id.vars = paste(condition,
                                                          collapse = "_"),
                                variable.name = "pathway")
gsvaResFlat$pathway <- factor(gsvaResFlat$pathway, levels = cls)
g1 <-  ggplot2::ggplot(gsvaResFlat, ggplot2::aes_string(
    x = "pathway", y = "value",
    color = "pathway")) +
    ggplot2::geom_boxplot(outlier.shape = NA) +
    ggpubr::stat_compare_means(method = "anova",  label = 'p.format', vjust = 0.75)+
    ggplot2::theme_bw()+
    ggplot2::facet_wrap(as.formula(paste("~",group)), scales = "free_y", ncol = nc) +
    ggplot2::theme(strip.text.x = ggplot2::element_text(size = 10))+
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, size = 8))+
    ggplot2::geom_hline(yintercept=0, linetype="dashed", color = "red")+
    ggplot2::theme(legend.position="none")+
    ggplot2::ylab("GSVA Score")
g1
}

```

```{r fig.width=15, fig.height=12}
gsva_boxplot(df=gsva_res_all, group = "seurat_clusters",nc = 7, cls = c('hknr_up', 'dys_up', 'cancer_up', 'hknr_dn', 'dys_dn', 'cancer_dn', 'hknr_diff', 'dys_diff', 'cancer_diff'))
gsva_boxplot(df=gsva_res_all, group = "treatment", nc = 4, cls = c('hknr_up', 'dys_up', 'cancer_up', 'hknr_dn', 'dys_dn', 'cancer_dn', 'hknr_diff', 'dys_diff', 'cancer_diff'))
gsva_boxplot(df=gsva_res_all, group = "fcm_cell_type", nc=4, cls = c('hknr_up', 'dys_up', 'cancer_up', 'hknr_dn', 'dys_dn', 'cancer_dn', 'hknr_diff', 'dys_diff', 'cancer_diff'))
gsva_boxplot(df=gsva_res_all, group = "fcm_cell_label", nc=4, cls = c('hknr_up', 'dys_up', 'cancer_up', 'hknr_dn', 'dys_dn', 'cancer_dn', 'hknr_diff', 'dys_diff', 'cancer_diff'))
```

```{r fig.width=15, fig.height=12}
gsva_boxplot(df=gsva_res_all_diff, group = "seurat_clusters",nc = 7, cls = c('hknr_diff', 'dys_diff', 'cancer_diff'))
gsva_boxplot(df=gsva_res_all_diff, group = "treatment", nc = 4, cls = c( 'hknr_diff', 'dys_diff', 'cancer_diff'))
gsva_boxplot(df=gsva_res_all_diff, group = "fcm_cell_type", nc=4, cls = c( 'hknr_diff', 'dys_diff', 'cancer_diff'))
gsva_boxplot(df=gsva_res_all_diff, group = "fcm_cell_label", nc=4, cls = c( 'hknr_diff', 'dys_diff', 'cancer_diff'))
```

```{r eval=FALSE}
#to export - NOTE: only the raw counts were exported to BT due to some issue with SCT normalised counts
rBCS::ExportSeurat(
  object = e7386_obj,
  bcs.path = file.path(PATH,"mmkhan_e7386/data/2023_06_13_e7386_seurat_filtered_main_final_BT.bcs"),
  batch.name = "treatment",
  unique.limit = 200,
  raw.rna = "RNA", # Assay that has raw count of RNA
  norm.rna = "RNA", # Assay that has normalized count of RNA  
  overwrite = TRUE # Overwrite existing bcs file
)
```
