---
title: "03.3.tcga_4nqo_e7386"
author: "M. Muzamil Khan"
date: "07/25/2023"
output:
  html_document:
    code_folding: hide
    theme: flatly
    toc: yes
    toc_float: true
  html_notebook:
    code_folding: hide
    theme: flatly
    toc: yes
    toc_float: true
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
nqoPATH <- "/restricted/projectnb/montilab-p/projects/oralcancer/4nqo/"
moscPATH <- "/restricted/projectnb/montilab-p/projects/oralcancer/4mosc1/"

library(Biobase)
library(tidyverse)
library(Seurat)
library(GSVA)
library(survminer)
library(survival)
library(ggplot2)
library(msigdbr)
library(adjustedCurves)
```


```{r eval=FALSE}
## Clean up eSet {.tabset .tabset-fade .tabset-pills}
## Add HPV status to eSet, tumor grade, and make sure gene names are hnsc symbols

require(Biobase)
require(tidyverse)
require(TCGAbiolinks)

eset <- readRDS("/restricted/projectnb/montilab-p/CBMrepositoryData/TCGA-GDC/esets_filtered/TCGA-HNSC_2020-03-22_DESeq2_log_filtered_eset.rds") # doesn't have tumor grade
eset_old <- readRDS("/restricted/projectnb/montilab-p/CBMrepositoryData/TCGA-GDC/esets_postprocessed/TCGA-HNSC_2020-03-22_wLegacyPData_DESeq2_log_eSet.rds") # gene names aren't hnsc symbols

# add back tumor grade from another eset
eset$hpv_status <- eset_old$hpv_status[match(eset_old$sample_id, eset$sample_id)]

# add FID to pData
eset$fid <- rownames(pData(eset))

# add back tumor grade from another eset
eset$tumor_grade <- eset_old$tumor_grade[match(eset_old$sample_id, eset$sample_id)]

pData(eset)

saveRDS(eset, file.path(nqoPATH, "mmkhan_e7386/data/2023_06_08_hnscc_eset_corrected.rds"))

```

```{r eval=FALSE}
### If needed use the code below to get hpv status from TCGABiolinks

# match sample id
rid2 <- substr(eset$sample_id, 1, nchar(eset$sample_id) - 4)

# Add hpv status
query <- GDCquery(
project = "TCGA-HNSC",
data.category = "Biospecimen",
data.type = "Biospecimen Supplement",
data.format = "BCR Biotab"
)

GDCdownload(query)

# get hpv status
hpv <- GDCprepare(query)$auxiliary_hnsc %>%
as.data.frame()
rownames(hpv) <- hpv$bcr_patient_barcode
eset$hpv_status <- hpv[rid2, "hpv_status"]
```

```{r eval=F}
data_lina <- readRDS("/restricted/projectnb/montilab-p/projects/oralcancer/4mosc1/data/TCGA-HNSC_edgeR_log_hpv_E7836_eSet.rds")
data_old <- readRDS("/restricted/projectnb/montilab-p/CBMrepositoryData/TCGA-GDC/esets_postprocessed/TCGA-OSCC_2020-03-22_wLegacyPData_DESeq2_log_eSet.rds")

eset_old <- readRDS("/restricted/projectnb/montilab-p/CBMrepositoryData/TCGA-GDC/esets_filtered/TCGA-HNSC_2020-03-22_DESeq2_log_filtered_eset.rds")


pData(data_old)

eset

pData(data_old)$sample_id %in% pData(data)$sample_id
pData(data)$sample_id
 
# Extract only HPV negative data
neg <- data_lina[ , data_lina$hpv_status == "neg"]
table(neg$hpv_status)

neg$stage <- data_old$tumor_stage[match(neg$sample_id, data_old$sample_id)]
neg$grade <- data_old$tumor_grade[match(neg$sample_id, data_old$sample_id)]

neg$molsub <- data_old$subtype_m_rna[match(neg$sample_id, data_old$sample_id)]
```

## GSVA TCGA analysis {.tabset .tabset-fade .tabset-pills}

Load Diffex Sigs

```{r}
`%ni%` <- Negate(`%in%`)
eset <- readRDS(file.path(nqoPATH, "mmkhan_e7386/data/2023_06_08_hnscc_eset_corrected.rds"))
eset <- eset[, eset$hpv_status == "negative"]
eset <- eset[, eset$sample_type != "Solid Tissue Normal"]
eset <- eset[, !eset$tumor_stage %in% c("not reported", "stage ivc")]
eset <- eset[, !eset$tumor_grade %in% c(NA)]
               
diff_list_nqo <- readRDS(file.path(nqoPATH, "mmkhan_e7386/results/06_08_4nqo_e7386_epi_sigs.rds"))
diff_list_nqo <- sapply(diff_list_nqo, toupper)


diff_list_e7386 <- c(diff_list_nqo)
diff_list_e7386 <- diff_list_e7386[c("nqo.vs.ctrl.up", "e7386_25.vs.nqo.up", "e7386_50.vs.nqo.up", "nqo.vs.ctrl.dn",  "e7386_25.vs.nqo.dn", "e7386_50.vs.nqo.dn")]
```

### GSVA

```{r}
nqo_gsva <- GSVA::gsva(exprs(eset), gset.idx.list = diff_list_e7386)
nqo_gsva_mod <- data.frame(t(nqo_gsva))

nqo_gsva_mod$nqo.diff <- nqo_gsva_mod$nqo.vs.ctrl.up - nqo_gsva_mod$nqo.vs.ctrl.dn
nqo_gsva_mod$e7386.25.diff <- nqo_gsva_mod$e7386_25.vs.nqo.dn - nqo_gsva_mod$e7386_25.vs.nqo.up
nqo_gsva_mod$e7386.50.diff <- nqo_gsva_mod$e7386_50.vs.nqo.dn - nqo_gsva_mod$e7386_50.vs.nqo.up
```


```{r}
gsvaplot <- function(df, condn){
  gsvaResT <- df
  condition <- condn
  cond <- as.list(gsvaResT[, condition ,drop = FALSE])
  gsvaResT[, paste(condition, collapse = "_")] <- cond
  gsvaResFlat <- reshape2::melt(gsvaResT, id.vars = paste(condition,
                                                          collapse = "_"),
                                variable.name = "pathway")
  g1 <-  ggplot2::ggplot(gsvaResFlat, ggplot2::aes_string(
    x = paste(condn), y = "value",
    color = "pathway")) +
    ggplot2::geom_boxplot(outlier.shape = NA) +
    ggpubr::stat_compare_means(method = "anova",  label = 'p.format', vjust = 0.75)+
    ggplot2::theme_bw()+
    ggplot2::facet_wrap(~paste(condn), scales = "free_y", ncol = 5 ) +
    ggplot2::theme(strip.text.x = ggplot2::element_text(size = 10))+
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, size = 8))+
    ggplot2::geom_hline(yintercept=0, linetype="dashed", color = "red")+
   #ggplot2::theme(legend.position="none")+
    ggplot2::ylab("GSVA Score")
g1
}
```

### Stage

```{r}
nqo_gsva_mod1 <- nqo_gsva_mod
nqo_gsva_mod1$stage <- eset$tumor_stage[match(eset$fid, rownames(nqo_gsva_mod))]
nqo_gsva_mod1 <- nqo_gsva_mod1[!nqo_gsva_mod1$stage %in% c("not reported", "stage ivc"), ]

nqo_gsva_mod1 <- nqo_gsva_mod1[complete.cases(nqo_gsva_mod1),]
gsvaplot(df = nqo_gsva_mod1, condn = "stage")

nqo_gsva_mod_sub1 <- nqo_gsva_mod1[, c(colnames(nqo_gsva_mod1)[grepl(x = colnames(nqo_gsva_mod1), pattern = "up")], "stage")]
gsvaplot(df = nqo_gsva_mod_sub1, condn = "stage")

nqo_gsva_mod_sub2 <- nqo_gsva_mod1[, c(colnames(nqo_gsva_mod1)[grepl(x = colnames(nqo_gsva_mod1), pattern = "dn")], "stage")]
gsvaplot(df = nqo_gsva_mod_sub2, condn = "stage")

nqo_gsva_mod_sub3 <- nqo_gsva_mod1[, c(colnames(nqo_gsva_mod1)[grepl(x = colnames(nqo_gsva_mod1), pattern = "diff")], "stage")]
gsvaplot(df = nqo_gsva_mod_sub3, condn = "stage")
```

### Grade

```{r}
nqo_gsva_mod2 <- nqo_gsva_mod
nqo_gsva_mod2$grade <- eset$tumor_grade[match(eset$fid, rownames(nqo_gsva_mod2))]

nqo_gsva_mod2 <- nqo_gsva_mod2[complete.cases(nqo_gsva_mod2),]
gsvaplot(df = nqo_gsva_mod2, condn = "grade")

nqo_gsva_mod_sub1 <- nqo_gsva_mod2[, c(colnames(nqo_gsva_mod2)[grepl(x = colnames(nqo_gsva_mod2), pattern = "up")], "grade")]
gsvaplot(df = nqo_gsva_mod_sub1, condn = "grade")

nqo_gsva_mod_sub2 <- nqo_gsva_mod2[, c(colnames(nqo_gsva_mod2)[grepl(x = colnames(nqo_gsva_mod2), pattern = "dn")], "grade")]
gsvaplot(df = nqo_gsva_mod_sub2, condn = "grade")

nqo_gsva_mod_sub3 <- nqo_gsva_mod2[, c(colnames(nqo_gsva_mod2)[grepl(x = colnames(nqo_gsva_mod2), pattern = "diff")], "grade")]
gsvaplot(df = nqo_gsva_mod_sub3, condn = "grade")
```

### Mol. Subtype

```{r}
nqo_gsva_mod3 <- nqo_gsva_mod
nqo_gsva_mod3$molsub <- eset$subtype_m_rna[match(eset$fid, rownames(nqo_gsva_mod3))]
nqo_gsva_mod3 <- nqo_gsva_mod3[complete.cases(nqo_gsva_mod3),]
gsvaplot(df = nqo_gsva_mod3, condn = "molsub")

nqo_gsva_mod_sub1 <- nqo_gsva_mod3[, c(colnames(nqo_gsva_mod3)[grepl(x = colnames(nqo_gsva_mod3), pattern = "up")], "molsub")]
gsvaplot(df = nqo_gsva_mod_sub1, condn = "molsub")

nqo_gsva_mod_sub2 <- nqo_gsva_mod3[, c(colnames(nqo_gsva_mod3)[grepl(x = colnames(nqo_gsva_mod3), pattern = "dn")], "molsub")]
gsvaplot(df = nqo_gsva_mod_sub2, condn = "molsub")

nqo_gsva_mod_sub3 <- nqo_gsva_mod3[, c(colnames(nqo_gsva_mod3)[grepl(x = colnames(nqo_gsva_mod3), pattern = "diff")], "molsub")]
gsvaplot(df = nqo_gsva_mod_sub3, condn = "molsub")
```

## Survival Analysis {.tabset .tabset-fade .tabset-pills}

```{r}
# use gsva to assign each sample a score for each signature.
# add back GSVA scores onto the eset

pData(eset) <- cbind(data.frame(pData(eset)[!names(pData(eset)) %in% names(nqo_gsva_mod)]), nqo_gsva_mod)

# add survival info to tcgadata
# pData(eset1) <- pData(eset1) %>% mutate(time = case_when(days_to_death != "--" ~ eset1$days_to_death, 
#                                                                days_to_death == "--" ~ eset1$days_to_last_follow_up))

pData(eset) <- pData(eset) %>% mutate(time = case_when(!is.na(days_to_death) ~ eset$days_to_death, 
                                                               is.na(days_to_death) ~ eset$days_to_last_follow_up))


```

```{r}
eset$vital_status1 <- ifelse(eset$vital_status == "Alive", 1, 2)

#5 year cutoff
pData(eset) <- pData(eset) %>% mutate(x5year = case_when(vital_status == "Alive" ~ as.character(eset$days_to_last_follow_up),
                                                                 (vital_status == "Alive" & as.numeric(days_to_last_follow_up) >= 1825.0) ~ as.character(1825),
                                                                 (vital_status == "Dead" & as.numeric(days_to_death) < 1825.0) ~ as.character(eset$time),
                                                                 (vital_status == "Dead" & as.numeric(days_to_death) >= 1825.0) ~ as.character(1825))
)
pData(eset) <- pData(eset) %>% mutate(vital_status5 = case_when(vital_status == "Alive" ~ 1,
                                                                        (vital_status == "Dead" & as.numeric(days_to_death) < 1825.0) ~ 2,
                                                                        (vital_status == "Dead" & as.numeric(days_to_death) >= 1825.0) ~ 1)
)

#function to add gsva scores to pData slot of eset 
addgsvadata <- function(obj) {
  for (sig in c("e7386_25.vs.nqo.dn", "e7386_50.vs.nqo.dn", "nqo.diff", "e7386.25.diff", "e7386.50.diff" )) {
    print(sig)
    threshold <- median(pData(obj)[, paste(sig)])
    print(threshold)
    #obj[[, paste0(sig,"val")]] <- pData(obj)[, paste(sig)]
    pData(obj)[, paste0(sig,"stat")] <- with(obj, ifelse(pData(obj)[, paste(sig)] <= threshold, "low", "high"))
  }
  return(obj)
}
#add gsva to eset Pdata
test <- addgsvadata(eset)

```

### KM

```{r}
#kaplan meier curves - less robust because doesn't take into account age
ggsurvplot(
  fit = survfit(Surv(as.numeric(test$time), test$vital_status1) ~ test$e7386_25.vs.nqo.dnstat),
  data = test, 
  xlab = "Days", 
  ylab = "H Overall survival probability",
  conf.int = TRUE,
  pval = TRUE)

ggsurvplot(
  fit = survfit(Surv(as.numeric(test$time), test$vital_status1) ~ test$e7386_50.vs.nqo.dnstat),
  data = test, 
  xlab = "Days", 
  ylab = "H Overall survival probability",
  conf.int = TRUE,
  pval = TRUE)


ggsurvplot(
  fit = survfit(Surv(as.numeric(test$time), test$vital_status1) ~ test$nqo.diffstat),
  data = test, 
  xlab = "Days", 
  ylab = "H Overall survival probability",
  conf.int = TRUE,
  pval = TRUE)

ggsurvplot(
  fit = survfit(Surv(as.numeric(test$time), test$vital_status1) ~ test$e7386.25.diffstat),
  data = test, 
  xlab = "Days", 
  ylab = "H Overall survival probability",
  conf.int = TRUE,
  pval = TRUE)

ggsurvplot(
  fit = survfit(Surv(as.numeric(test$time), test$vital_status1) ~ test$e7386.50.diffstat),
  data = test, 
  xlab = "Days", 
  ylab = "H Overall survival probability",
  conf.int = TRUE,
  pval = TRUE)



cox_mod <- coxph(Surv(as.numeric(test$time), test$vital_status1) ~ test$e7386.50.diffstat,
                 data=test, x=TRUE)

test$e7386.50.diffstat <- factor(test$e7386.50.diffstat)
test$vital_status1 <- as.numeric(test$vital_status1)
# using g-computation with confidence intervals
adjsurv <- adjustedsurv(data=pData(test),
                        variable="e7386.50.diffstat",
                        ev_time="time",
                        event="vital_status1",
                        method="direct",
                        outcome_model=cox_mod,
                        conf_int=TRUE,
                        bootstrap=FALSE)
plot(adjsurv, conf_int=TRUE)

```

```{r}
## example
library(adjustedCurves)
library(survival)
library(riskRegression)
library(pammtools)

# just used to make the example reproducible
set.seed(42)

# simulate some example data
sim_dat <- sim_confounded_surv(n=50, max_t=1.2)
sim_dat$group <- as.factor(sim_dat$group)

# outcome model
cox_mod <- coxph(Surv(time, event) ~ x1 + x2 + x4 + x5 + group,
                 data=sim_dat, x=TRUE)

# using g-computation with confidence intervals
adjsurv <- adjustedsurv(data=sim_dat,
                        variable="group",
                        ev_time="time",
                        event="event",
                        method="direct",
                        outcome_model=cox_mod,
                        conf_int=TRUE,
                        bootstrap=FALSE)
plot(adjsurv, conf_int=TRUE, )

```


### CoxPH {.tabset .tabset-fade .tabset-pills}

#### Status + Age {.tabset .tabset-fade .tabset-pills}

```{r}
#add age metadata in years
test$age <- as.numeric(test$age_at_diagnosis)/365
#add age stratification -- this is based off cuttoffs Anthony S wanted to use, you may want to skip or change this.
threshold1 <- 80
theshold2 <- 30
test$agestat <- with(test, ifelse(test$age <= threshold1, "young", "old"))
test$agestat1 <- with(test, ifelse(test$age >= theshold2 & test$agestat == 'young', "mid", test$agestat))

test$agestat1 <- factor(test$agestat1, levels = c("young", "mid", "old"))
```

##### E7386_25 dn

```{r}
#with age as covariate and as stratifying variable
test$e7386_25.vs.nqo.dnstat <- factor(test$e7386_25.vs.nqo.dnstat, levels = c("low", "high"))
coxphres <- coxph(formula = Surv(as.numeric(test$time), test$vital_status1) ~ test$e7386_25.vs.nqo.dnstat + test$age + strata(test$agestat1))
summary(coxphres)
```

##### E7386_50 dn

```{r}
#with age as covariate and as stratifying variable
test$e7386_50.vs.nqo.dnstat <- factor(test$e7386_50.vs.nqo.dnstat, levels = c("low", "high"))
coxphres <- coxph(formula = Surv(as.numeric(test$time), test$vital_status1) ~ test$e7386_50.vs.nqo.dnstat + test$age + strata(test$agestat1))
summary(coxphres)
```


##### 4NQO diff 

```{r}
#with age as covariate and as stratifying variable
coxphres <- coxph(formula = Surv(as.numeric(test$time), test$vital_status1) ~ test$nqo.diffstat + test$age + strata(test$agestat1))
summary(coxphres)
```

##### 4NQO + E7386_25

```{r}
coxphres <- coxph(formula = Surv(as.numeric(test$time), test$vital_status1) ~ test$e7386.25.diffstat + test$age + strata(test$agestat1))
summary(coxphres)
```

##### 4NQO + E7386_50

```{r}
coxphres <- coxph(formula = Surv(as.numeric(test$time), test$vital_status1) ~ test$e7386.50.diffstat + test$age + strata(test$agestat1))
summary(coxphres)
```

#### Status {.tabset .tabset-fade .tabset-pills}

##### E7386_25 dn

```{r}
coxphres2 <- coxph(formula = Surv(as.numeric(test$time), test$vital_status1) ~ test$e7386_25.vs.nqo.dnstat + strata(test$agestat1))
summary(coxphres2)
```

##### E7386_50 dn

```{r}
coxphres2 <- coxph(formula = Surv(as.numeric(test$time), test$vital_status1) ~ test$e7386_50.vs.nqo.dnstat + strata(test$agestat1))
summary(coxphres2)
```


##### 4NQO

```{r}
coxphres2 <- coxph(formula = Surv(as.numeric(test$time), test$vital_status1) ~ test$nqo.diffstat + strata(test$agestat1))
summary(coxphres2)
```

##### 4NQO + E7386_25

```{r}
coxphres2 <- coxph(formula = Surv(as.numeric(test$time), test$vital_status1) ~ test$e7386.25.diffstat + strata(test$agestat1))
summary(coxphres2)
```

##### 4NQO + E7386_50

```{r}
coxphres2 <- coxph(formula = Surv(as.numeric(test$time), test$vital_status1) ~ test$e7386.50.diffstat + strata(test$agestat1))
summary(coxphres2)
```

# Stress signatures

unique to cluster 4 and 11

```{r}
fcm_stress <- read.csv(file = file.path(nqoPATH, "mmkhan_e7386/results/stress_unique_genes_per_clust.csv"))

fcm_stress_list <- apply(fcm_stress, 2, function(x) list(x))
fcm_stress_list$X <- NULL
#remove NAs
fcm_stress_list_mod <- lapply(fcm_stress_list, function(x) unique(x[[1]])[!is.na(unique(x[[1]]))])


```
```{r}
gsva_res_stress_411 <- GSVA::gsva(exprs(eset), gset.idx.list = fcm_stress_list_mod)
gsva_res_stress_411_mod <- data.frame(t(gsva_res_stress_411))
```


```{r}
# use gsva to assign each sample a score for each signature.
# add back GSVA scores onto the eset

pData(eset) <- cbind(data.frame(pData(eset)[!names(pData(eset)) %in% names(gsva_res_stress_411_mod)]), gsva_res_stress_411_mod)

```

```{r}
# eset$vital_status1 <- ifelse(eset$vital_status == "Alive", 1, 2)
# 
# #5 year cutoff
# pData(eset) <- pData(eset) %>% mutate(x5year = case_when(vital_status == "Alive" ~ as.character(eset$days_to_last_follow_up),
#                                                                  (vital_status == "Alive" & as.numeric(days_to_last_follow_up) >= 1825.0) ~ as.character(1825),
#                                                                  (vital_status == "Dead" & as.numeric(days_to_death) < 1825.0) ~ as.character(eset$time),
#                                                                  (vital_status == "Dead" & as.numeric(days_to_death) >= 1825.0) ~ as.character(1825))
# )
# pData(eset) <- pData(eset) %>% mutate(vital_status5 = case_when(vital_status == "Alive" ~ 1,
#                                                                         (vital_status == "Dead" & as.numeric(days_to_death) < 1825.0) ~ 2,
#                                                                         (vital_status == "Dead" & as.numeric(days_to_death) >= 1825.0) ~ 1)
# )

#function to add gsva scores to pData slot of eset 
addgsvadata <- function(obj) {
  for (sig in c("clust4_w11", "clust11_w4", "clust11_wo4", "clust4_wo11")) {
    print(sig)
    threshold <- median(pData(obj)[, paste(sig)])
    print(threshold)
    #obj[[, paste0(sig,"val")]] <- pData(obj)[, paste(sig)]
    pData(obj)[, paste0(sig,"stat")] <- with(obj, ifelse(pData(obj)[, paste(sig)] <= threshold, "low", "high"))
  }
  return(obj)
}
#add gsva to eset Pdata
test <- addgsvadata(eset)

```



### KM

```{r}
#kaplan meier curves - less robust because doesn't take into account age
ggsurvplot(
  fit = survfit(Surv(as.numeric(test$time), test$vital_status1) ~ test$clust4_wo11stat),
  data = test, 
  xlab = "Days", 
  ylab = "H Overall survival probability",
  conf.int = TRUE,
  pval = TRUE)

ggsurvplot(
  fit = survfit(Surv(as.numeric(test$time), test$vital_status1) ~ test$clust11_wo4stat),
  data = test, 
  xlab = "Days", 
  ylab = "H Overall survival probability",
  conf.int = TRUE,
  pval = TRUE)

ggsurvplot(
  fit = survfit(Surv(as.numeric(test$time), test$vital_status1) ~ test$clust4_w11stat),
  data = test, 
  xlab = "Days", 
  ylab = "H Overall survival probability",
  conf.int = TRUE,
  pval = TRUE)

ggsurvplot(
  fit = survfit(Surv(as.numeric(test$time), test$vital_status1) ~ test$clust11_w4stat),
  data = test, 
  xlab = "Days", 
  ylab = "H Overall survival probability",
  conf.int = TRUE,
  pval = TRUE)

```


### CoxPH {.tabset .tabset-fade .tabset-pills}

#### Status + Age {.tabset .tabset-fade .tabset-pills}

```{r}
#add age metadata in years
test$age <- as.numeric(test$age_at_diagnosis)/365
#add age stratification -- this is based off cuttoffs Anthony S wanted to use, you may want to skip or change this.
threshold1 <- 80
theshold2 <- 30
test$agestat <- with(test, ifelse(test$age <= threshold1, "young", "old"))
test$agestat1 <- with(test, ifelse(test$age >= theshold2 & test$agestat == 'young', "mid", test$agestat))
```

##### 4

```{r}
#with age as covariate and as stratifying variable
coxphres <- coxph(formula = Surv(as.numeric(test$time), test$vital_status1) ~ test$clust4_wo11stat + test$age + strata(test$agestat1))
summary(coxphres)
```

##### 11

```{r}
coxphres <- coxph(formula = Surv(as.numeric(test$time), test$vital_status1) ~ test$clust11_wo4stat + test$age + strata(test$agestat1))
summary(coxphres)
```

##### 4

```{r}
coxphres <- coxph(formula = Surv(as.numeric(test$time), test$vital_status1) ~ test$clust4_w11stat + test$age + strata(test$agestat1))
summary(coxphres)
```


##### 11

```{r}
coxphres <- coxph(formula = Surv(as.numeric(test$time), test$vital_status1) ~ test$clust11_w4stat + test$age + strata(test$agestat1))
summary(coxphres)
```


#### Status {.tabset .tabset-fade .tabset-pills}

##### 4

```{r}
coxphres2 <- coxph(formula = Surv(as.numeric(test$time), test$vital_status1) ~ test$clust4_wo11stat + strata(test$agestat1))
summary(coxphres2)
```

##### 11

```{r}
coxphres2 <- coxph(formula = Surv(as.numeric(test$time), test$vital_status1) ~ test$clust11_wo4stat + strata(test$agestat1))
summary(coxphres2)
```

##### 4 w/ 11

```{r}
coxphres2 <- coxph(formula = Surv(as.numeric(test$time), test$vital_status1) ~ test$clust4_w11stat + strata(test$agestat1))
summary(coxphres2)
```

##### 11 wo/ 4

```{r}
coxphres2 <- coxph(formula = Surv(as.numeric(test$time), test$vital_status1) ~ test$clust11_w4stat + strata(test$agestat1))
summary(coxphres2)
```



