---
title: "Connectivity Map Querying"
subtitle: "ICG-001 and E7386 signatures"
author: "Stefano Monti"
output:
  html_document:
    theme: united
    code_folding: show
    css: style.css
    toc: yes
---

```{r global, echo=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE,fig.path='../results/figures/')
```

In this document, we (re)derive the ICG-001 signatures. Furthermore, we aim to reconcile the results with the legacy results published in the Genome Medicine paper. This is just an "excerpt" from the original file (w/o the '.updated' suffix).

```{r settings}
library(OmicSignature)

PATH <- file.path(Sys.getenv("MLAB"),"projects/oralcancer/bcat_cbp_inhibition/4nqo/smonti")
SPATH <- file.path(Sys.getenv("MLAB"), "projects/sigRepo")
options(box.path = file.path(Sys.getenv("CBMGIT"), "MLscripts"))
box::use(R/MLmisc)

## control variables
do_save <- TRUE
```

# Load the data
```{r load.data}
## see the legacy file signature.icg001_hnsc.Rmd for the derivation of these data
icg001_hsc3 <- readRDS(file = file.path(SPATH, "archive/results/icg001_hsc3_hnsc_diffanal.RDS")) |>
  dplyr::rename(p_value = "p")
icg001_cal27 <- readRDS(file = file.path(SPATH, "archive/results/icg001_cal27_hnsc_diffanal.RDS")) |>
  dplyr::rename(p_value = "p")
e7386_hsc3 <- readRDS(file = file.path(SPATH, "archive/results/e7386_hsc3_hnsc_diffanal.RDS")) |>
  dplyr::rename(p_value = "p")
e7386_cal27 <- readRDS(file = file.path(SPATH, "archive/results/e7386_cal27_hnsc_diffanal.RDS")) |>
  dplyr::rename(p_value = "p")
```

```{r check}
icg001_sigs <- readRDS(file.path(SPATH, "archive/results/icg001_hnsc_signatures.RDS"))
e7386_sigs <- readRDS(file.path(SPATH, "archive/results/e7386_hnsc_signatures.RDS"))
data.frame(e7386 = lengths(e7386_sigs), 
           icg001 = lengths(icg001_sigs))
```

# Define Relevant Differential Signatures {.tabset}

```{r icg001.signatures}
icg001_hsc3_flt <- tidyr::drop_na(icg001_hsc3, FDR.q.filtered, fold.change, Symbol)
icg001_cal27_flt <- tidyr::drop_na(icg001_cal27, FDR.q.filtered, fold.change, Symbol)
e7386_hsc3_flt <- tidyr::drop_na(e7386_hsc3, FDR.q.filtered, fold.change, Symbol)
e7386_cal27_flt <- tidyr::drop_na(e7386_cal27, FDR.q.filtered, fold.change, Symbol)

maxQ <- 0.01
minFC <- 1.5
```

## ICG-001
For consistency with past results, we use the signatures derived based on rounded values.
```{r icg001.signatures}
icg001sigRound <- list(
  hsc3.up = icg001_hsc3_flt |>
    dplyr::filter(signif(FDR.q.filtered, 2) <= maxQ, round(fold.change, 1) >= +minFC) |>
    dplyr::pull(Symbol),
  hsc3.dn = icg001_hsc3_flt |>
    dplyr::filter(signif(FDR.q.filtered, 2) <= maxQ, round(fold.change, 1) <= -minFC) |>
    dplyr::pull(Symbol),
  cal27.up = icg001_cal27_flt |>
    dplyr::filter(signif(FDR.q.filtered, 2) <= maxQ, round(fold.change, 1) >= +minFC) |>
    dplyr::pull(Symbol),
  cal27.dn = icg001_cal27_flt |>
    dplyr::filter(signif(FDR.q.filtered, 2) <= maxQ, round(fold.change, 1) <= -minFC) |>
    dplyr::pull(Symbol)
)
icg001sigRound <- c(
  icg001sigRound,
  list(comb.up=intersect(icg001sigRound$hsc3.up,icg001sigRound$cal27.up),
       comb.dn=intersect(icg001sigRound$hsc3.dn,icg001sigRound$cal27.dn))
)
names(icg001sigRound) <- paste("icg001",names(icg001sigRound),sep=".")
```

## E7386
```{r e7386.signatures}
e7386sigRound <- list(
  hsc3.up = e7386_hsc3_flt |>
    dplyr::filter(signif(FDR.q.filtered, 2) <= maxQ, round(fold.change, 1) >= +minFC) |>
    dplyr::pull(Symbol),
  hsc3.dn = e7386_hsc3_flt |>
    dplyr::filter(signif(FDR.q.filtered, 2) <= maxQ, round(fold.change, 1) <= -minFC) |>
    dplyr::pull(Symbol),
  cal27.up = e7386_cal27_flt |>
    dplyr::filter(signif(FDR.q.filtered, 2) <= maxQ, round(fold.change, 1) >= +minFC) |>
    dplyr::pull(Symbol),
  cal27.dn = e7386_cal27_flt |>
    dplyr::filter(signif(FDR.q.filtered, 2) <= maxQ, round(fold.change, 1) <= -minFC) |>
    dplyr::pull(Symbol)
)
e7386sigRound <- c(
  e7386sigRound,
  list(comb.up=intersect(e7386sigRound$hsc3.up,e7386sigRound$cal27.up),
       comb.dn=intersect(e7386sigRound$hsc3.dn,e7386sigRound$cal27.dn))
)
names(e7386sigRound) <- paste("e7386",names(e7386sigRound),sep=".")
```

# {-}

```{r}
data.frame(
  e7386 = lengths(e7386sigRound),
  icg001 = lengths(icg001sigRound)
)
```

# CMap Query
```{r} 
e7386_cmap_query <- e7386sigRound[grep("hsc3", names(e7386sigRound))]
icg001_cmap_query <- icg001sigRound[grep("hsc3", names(icg001sigRound))]

e7386_cmap_query_flt <- lapply(
  e7386_cmap_query,
  \(x) x[!grepl("\\bLOC", x, perl = TRUE)]
)
icg001_cmap_query_flt <- lapply(
  icg001_cmap_query,
  \(x) x[!grepl("\\bLOC", x, perl = TRUE)]
)
data.frame(
  unfiltered = lengths(c(e7386_cmap_query, icg001_cmap_query)),
  filtered = lengths(c(e7386_cmap_query_flt, icg001_cmap_query_flt))
)
cmap_query_df <- list(
  E7386_UP = e7386_cmap_query_flt$e7386.hsc3.up,
  E7386_DN = e7386_cmap_query_flt$e7386.hsc3.dn,
  ICG_001_UP = icg001_cmap_query_flt$icg001.hsc3.up,
  ICG_001_DN = icg001_cmap_query_flt$icg001.hsc3.dn
) |> MLmisc$list2DF()
```

## API call
```{r eval=FALSE, echo=FALSE}
## THIS DOESN'T WORK
cmap_api_key <- Sys.getenv("CLUE_API_KEY")
if (!nzchar(cmap_api_key)) {
  cmap_api_key <- Sys.getenv("CMAP_API_KEY")
}
if (!nzchar(cmap_api_key)) {
  stop("Set CLUE_API_KEY (or CMAP_API_KEY) to your clue.io API key before submitting CMap queries.")
}

cmap_api_url <- Sys.getenv("CLUE_API_URL", unset = "https://api.clue.io/api/jobs")

as_cmap_gmt <- function(query, direction) {
  genes <- query[grep(paste0("\\.", direction, "$"), names(query))][[1]]
  paste(c("TAG", "", unique(stats::na.omit(genes))), collapse = "\t")
}

submit_cmap_query <- function(query, name) {
  body <- list(
    tool_id = "sig_gutc_tool",
    data_type = "L1000",
    name = name,
    dataset = "Touchstone",
    ignoreWarnings = TRUE,
    "uptag-cmapfile" = as_cmap_gmt(query, "up"),
    "dntag-cmapfile" = as_cmap_gmt(query, "dn")
  )

  response <- httr::POST(
    url = cmap_api_url,
    httr::add_headers(user_key = cmap_api_key, Accept = "application/json"),
    httr::content_type_json(),
    body = body,
    encode = "json"
  )
  httr::stop_for_status(response)
  httr::content(response, as = "parsed", type = "application/json")
}

# submit E7386 query through the CMap API
e7386_cmap_query_res <- submit_cmap_query(
  query = e7386_cmap_query,
  name = "E7386 HSC3 signature"
)

# submit ICG001 query through the CMap API
icg001_cmap_query_res <- submit_cmap_query(
  query = icg001_cmap_query,
  name = "ICG001 HSC3 signature"
)
```


# Save
```{r save, echo=do_save, eval=do_save}
saveRDS(e7386_cmap_query_flt, file =
          file.path(PATH, "workdir/e7386_cmap_query.Rds"))
saveRDS(icg001_cmap_query_flt,
        file.path(PATH, "workdir/icg001_cmap_query.Rds"))
openxlsx::write.xlsx(cmap_query_df, file = file.path(PATH, "results/bcat_cbp_cmap_query.xlsx"))
```
