---
title: "LINCS signatureSearch Querying"
subtitle: "ICG-001 and E7386 signatures"
author: "Stefano Monti"
output:
  html_document:
    theme: united
    code_folding: show
    css: style.css
    toc: yes
---

```{r global, echo=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, fig.path='../results/figures/')
```

E7386 and ICG-001 are Wnt/beta-catenin/CBP pathway inhibitors. Here we use their HSC3 transcriptional signatures as LINCS queries to ask two related questions:

1. Which LINCS perturbations induce expression programs that are similar to, or opposite from, the E7386 and ICG-001 signatures?
2. How similar are E7386 and ICG-001 to each other when represented by their LINCS perturbation-similarity profiles?

Positive LINCS connectivity scores indicate that a perturbation resembles the query signature, whereas negative scores indicate an inverse relationship. We focus on `NCS`, the normalized connectivity score for perturbation/cell-line signatures, and `NCSct`, the LINCS cell-line summarized score that gives one value per perturbation.

```{r settings}
# Load analysis packages and configure project paths.
library(OmicSignature)
library(signatureSearch)

PATH <- file.path(Sys.getenv("MLAB"),"projects/oralcancer/bcat_cbp_inhibition/4nqo/smonti/")
options(box.path = file.path(Sys.getenv("CBMGIT"), "MLscripts"))
box::use(R/MLmisc)

## control variables
do_save <- TRUE
use_cached <- TRUE
```

# Load signatures

The query signatures were generated in the previous step. Each object contains ordered up- and down-regulated gene symbols for the E7386 or ICG-001 response in HSC3 cells.

```{r load.signatures}
# Read filtered HSC3 up/down gene sets saved by the previous notebook.
e7386_query <- readRDS(file = file.path(PATH, "workdir/e7386_cmap_query.Rds"))
icg001_query <- readRDS(file.path(PATH, "workdir/icg001_cmap_query.Rds"))
```

# Run LINCS `signatureSearch`

We query the LINCS L1000 resource exposed through `signatureSearchData`. The LINCS database stores moderated z-score signatures and `gess_lincs()` returns `WTCS`, `NCS`, and `NCSct` scores.

## Load LINCS Database

```{r eval=!use_cached}
# Retrieve the pre-built LINCS L1000 HDF5 reference database.
library(ExperimentHub)
library(rhdf5)
eh <- ExperimentHub::ExperimentHub()

# LINCS L1000: moderated z-score signatures.
lincs <- eh[["EH3226"]]
lincs_gene_ids <- rhdf5::h5read(lincs, "rownames")
```

```{r eval=use_cached, echo=FALSE}
# Use pre-computed LINCS query and result objects.
e7386_lincs_query_sig <- readRDS(file.path(PATH, "workdir/e7386_signatureSearch_lincs_query.Rds"))
icg001_lincs_query_sig <- readRDS(file.path(PATH, "workdir/icg001_signatureSearch_lincs_query.Rds"))
e7386_lincs_query_res <- readRDS(file.path(PATH, "workdir/e7386_signatureSearch_lincs_res.Rds"))
icg001_lincs_query_res <- readRDS(file.path(PATH, "workdir/icg001_signatureSearch_lincs_res.Rds"))
e7386_lincs_query_tbl <- readRDS(file.path(PATH, "workdir/e7386_signatureSearch_lincs_tbl.Rds"))
icg001_lincs_query_tbl <- readRDS(file.path(PATH, "workdir/icg001_signatureSearch_lincs_tbl.Rds"))
e7386_lincs_ncsct_tbl <- readRDS(file.path(PATH, "workdir/e7386_signatureSearch_lincs_ncsct_tbl.Rds"))
icg001_lincs_ncsct_tbl <- readRDS(file.path(PATH, "workdir/icg001_signatureSearch_lincs_ncsct_tbl.Rds"))
```

## Build LINCS Queries

For each inhibitor, we build a bidirectional query signature with exactly 150 up and 150 down Entrez IDs that are present in the LINCS database. We start with a larger pool of symbols because some symbols do not map to Entrez IDs or are absent from LINCS.

```{r eval=!use_cached}
# Use the top 150 mapped LINCS genes per direction.
top_n_markers <- 150
mapping_pool_size <- 300

map_top_entrez <- function(symbols, ref_gene_ids, n = 150, pool_size = 300) {
  # Over-select symbols first so unmapped or unavailable genes do not shrink the query.
  symbols <- symbols |>
    unique() |>
    head(pool_size)

  # Convert HGNC symbols to Entrez IDs required by signatureSearch.
  entrez <- AnnotationDbi::mapIds(
    org.Hs.eg.db::org.Hs.eg.db,
    keys = symbols,
    column = "ENTREZID",
    keytype = "SYMBOL",
    multiVals = "first"
  )

  # Keep only unique Entrez IDs represented in the LINCS reference database.
  entrez <- unique(unname(stats::na.omit(entrez)))
  entrez <- entrez[entrez %in% ref_gene_ids]
  if (length(entrez) < n) {
    stop("Only ", length(entrez), " symbols mapped to Entrez IDs in LINCS; need ", n, ".")
  }
  head(entrez, n)
}

top_signature_search_query <- function(query, ref_gene_ids, n = 150, pool_size = 300) {
  # Split the saved query object into up- and down-regulated symbol vectors.
  up_symbols <- query[grep("\\.up$", names(query))][[1]] |>
    unique()
  down_symbols <- query[grep("\\.dn$", names(query))][[1]] |>
    unique()

  # Return the list structure expected by qSig for the LINCS method.
  list(
    upset = map_top_entrez(up_symbols, ref_gene_ids = ref_gene_ids, n = n, pool_size = pool_size),
    downset = map_top_entrez(down_symbols, ref_gene_ids = ref_gene_ids, n = n, pool_size = pool_size)
  )
}

# Construct LINCS-compatible queries for both inhibitors.
e7386_lincs_query_sig <- top_signature_search_query(
  e7386_query,
  ref_gene_ids = lincs_gene_ids,
  n = top_n_markers,
  pool_size = mapping_pool_size
)
icg001_lincs_query_sig <- top_signature_search_query(
  icg001_query,
  ref_gene_ids = lincs_gene_ids,
  n = top_n_markers,
  pool_size = mapping_pool_size
)

data.frame(
  # Confirm that each LINCS query uses 150 genes per direction.
  compound = c("E7386", "ICG001"),
  up = c(length(e7386_lincs_query_sig$upset), length(icg001_lincs_query_sig$upset)),
  down = c(length(e7386_lincs_query_sig$downset), length(icg001_lincs_query_sig$downset))
)
```

## LINCS Search

The LINCS search compares each inhibitor query against L1000 perturbation signatures. `NCS` is the normalized connectivity score for each perturbation/cell-line signature. `NCSct` summarizes scores across cell lines for each perturbation.

```{r eval=!use_cached}
run_lincs_signature_search <- function(query, refdb = lincs, ref_gene_ids, n = 150, pool_size = 300) {
  # Build a LINCS-compatible qSig and run the LINCS weighted connectivity search.
  qsig <- qSig(
    query = top_signature_search_query(query, ref_gene_ids = ref_gene_ids, n = n, pool_size = pool_size),
    gess_method = "LINCS",
    refdb = refdb
  )

  # tau=FALSE avoids the more expensive Tau calculation; NCSct is still returned.
  gess_lincs(qSig = qsig, sortby = "NCSct", tau = FALSE)
}

# Run LINCS searches for E7386 and ICG-001.
e7386_lincs_query_res <- run_lincs_signature_search(
  query = e7386_query,
  refdb = lincs,
  ref_gene_ids = lincs_gene_ids,
  n = top_n_markers,
  pool_size = mapping_pool_size
)
icg001_lincs_query_res <- run_lincs_signature_search(
  query = icg001_query,
  refdb = lincs,
  ref_gene_ids = lincs_gene_ids,
  n = top_n_markers,
  pool_size = mapping_pool_size
)

# Extract per-signature LINCS results.
e7386_lincs_query_tbl <- result(e7386_lincs_query_res)
icg001_lincs_query_tbl <- result(icg001_lincs_query_res)

# Keep one row per perturbation using the cell-line summarized NCSct score.
e7386_lincs_ncsct_tbl <- e7386_lincs_query_tbl |>
  dplyr::distinct(pert, type, NCSct, .keep_all = TRUE) |>
  dplyr::arrange(desc(abs(NCSct)))
icg001_lincs_ncsct_tbl <- icg001_lincs_query_tbl |>
  dplyr::distinct(pert, type, NCSct, .keep_all = TRUE) |>
  dplyr::arrange(desc(abs(NCSct)))
```

```{r}
# Preview the strongest LINCS perturbation-level matches for each inhibitor.
list(
  E7386 = head(e7386_lincs_ncsct_tbl[, c("pert", "cell", "type", "NCS", "NCSct")]),
  ICG001 = head(icg001_lincs_ncsct_tbl[, c("pert", "cell", "type", "NCS", "NCSct")])
)
```

## Similarity Between E7386 and ICG-001 Profiles

Beyond ranking individual hits, we compare E7386 and ICG-001 by correlating their LINCS perturbation-similarity profiles. If the two inhibitors have similar mechanisms or downstream consequences, then perturbations that score highly for E7386 should tend to score highly for ICG-001 as well.

```{r}
compare_similarity_profiles <- function(tbl_a, tbl_b, score_col, by_cols) {
  # Join two result tables on shared perturbation identifiers and compare scores.
  joined <- dplyr::inner_join(
    tbl_a |> dplyr::select(dplyr::all_of(c(by_cols, score_col))),
    tbl_b |> dplyr::select(dplyr::all_of(c(by_cols, score_col))),
    by = by_cols,
    suffix = c("_e7386", "_icg001")
  )
  score_e7386 <- paste0(score_col, "_e7386")
  score_icg001 <- paste0(score_col, "_icg001")
  df <- data.frame(
    score = score_col,
    n = nrow(joined),
    pearson = stats::cor(joined[[score_e7386]], joined[[score_icg001]],
                         method = "pearson", use = "complete.obs"),
    spearman = stats::cor(joined[[score_e7386]], joined[[score_icg001]],
                          method = "spearman", use = "complete.obs")
  )
  p1 <- ggplot2::ggplot(joined, ggplot2::aes(
    x = .data[[score_e7386]], y = .data[[score_icg001]]
  )) +
#    ggplot2::geom_point(ggplot2::aes(col = cell)) +
    ggplot2::geom_point() +
    ggplot2::geom_smooth(method = lm)

  return(list(comparison = df, plot = p1))
}

# Compare full perturbation/cell-line profiles using NCS.
lincs_profile_similarity <- compare_similarity_profiles(
  e7386_lincs_query_tbl,
  icg001_lincs_query_tbl,
  score_col = "NCS",
  by_cols = c("pert", "cell", "type")
)

# Compare perturbation-level profiles using NCSct.
lincs_ncsct_profile_similarity <- compare_similarity_profiles(
  e7386_lincs_ncsct_tbl,
  icg001_lincs_ncsct_tbl,
  score_col = "NCSct",
  by_cols = c("pert", "type")
)

dplyr::bind_rows(
  cbind(database = "LINCS", level = "perturbation-cell", lincs_profile_similarity$comparison),
  cbind(database = "LINCS", level = "perturbation", lincs_ncsct_profile_similarity$comparison)
)

lincs_profile_similarity$plot
lincs_ncsct_profile_similarity$plot
```

## Mechanism-of-Action Summaries

The LINCS results include `MOAss`, a semicolon-delimited mechanism-of-action annotation for some perturbagens. To identify mechanisms that are collectively similar or opposite to each inhibitor signature, we remove missing MOA annotations, split multi-MOA entries, and average `NCSct` by MOA.

```{r}
summarize_ncsct_by_moa <- function(tbl) {
  tbl |>
    # Restrict to perturbagens with mechanism-of-action annotations.
    tidyr::drop_na(MOAss) |>
    dplyr::filter(MOAss != "") |>
    # Expand semicolon-delimited MOA annotations to one MOA per row.
    tidyr::separate_longer_delim(MOAss, delim = ";") |>
    dplyr::mutate(MOAss = trimws(MOAss)) |>
    dplyr::filter(MOAss != "") |>
    dplyr::group_by(MOAss) |>
    dplyr::summarize(
      n_perturbations = dplyr::n_distinct(pert),
      mean_NCSct = mean(NCSct, na.rm = TRUE),
      median_NCSct = median(NCSct, na.rm = TRUE),
      max_abs_NCSct = NCSct[which.max(abs(NCSct))],
      .groups = "drop"
    ) |>
    dplyr::arrange(desc(abs(mean_NCSct)))
}
e7386_lincs_moa_ncsct_tbl <- summarize_ncsct_by_moa(e7386_lincs_ncsct_tbl)
icg001_lincs_moa_ncsct_tbl <- summarize_ncsct_by_moa(icg001_lincs_ncsct_tbl)

list(
  E7386 = head(e7386_lincs_moa_ncsct_tbl),
  ICG001 = head(icg001_lincs_moa_ncsct_tbl)
)
```

# Save

We save the LINCS query signatures, full result objects, result tables, perturbation-level summaries, and E7386-vs-ICG001 profile similarity summaries for downstream inspection without rerunning the full search.

```{r save, echo=do_save, eval=do_save}
# Save LINCS query signatures and search outputs.
saveRDS(e7386_lincs_query_sig,
        file = file.path(PATH, "workdir/e7386_signatureSearch_lincs_query.Rds"))
saveRDS(icg001_lincs_query_sig,
        file = file.path(PATH, "workdir/icg001_signatureSearch_lincs_query.Rds"))
saveRDS(e7386_lincs_query_res,
        file = file.path(PATH, "workdir/e7386_signatureSearch_lincs_res.Rds"))
saveRDS(icg001_lincs_query_res,
        file = file.path(PATH, "workdir/icg001_signatureSearch_lincs_res.Rds"))
saveRDS(e7386_lincs_query_tbl,
        file = file.path(PATH, "workdir/e7386_signatureSearch_lincs_tbl.Rds"))
saveRDS(icg001_lincs_query_tbl,
        file = file.path(PATH, "workdir/icg001_signatureSearch_lincs_tbl.Rds"))
saveRDS(e7386_lincs_ncsct_tbl,
        file = file.path(PATH, "workdir/e7386_signatureSearch_lincs_ncsct_tbl.Rds"))
saveRDS(icg001_lincs_ncsct_tbl,
        file = file.path(PATH, "workdir/icg001_signatureSearch_lincs_ncsct_tbl.Rds"))
saveRDS(e7386_lincs_moa_ncsct_tbl,
        file = file.path(PATH, "workdir/e7386_signatureSearch_lincs_moa_ncsct_tbl.Rds"))
saveRDS(icg001_lincs_moa_ncsct_tbl,
        file = file.path(PATH, "workdir/icg001_signatureSearch_lincs_moa_ncsct_tbl.Rds"))

# Save profile-level similarity summaries comparing E7386 and ICG-001.
saveRDS(lincs_profile_similarity,
        file = file.path(PATH, "workdir/e7386_icg001_lincs_profile_similarity.Rds"))
saveRDS(lincs_ncsct_profile_similarity,
        file = file.path(PATH, "workdir/e7386_icg001_lincs_ncsct_profile_similarity.Rds"))
```
