---
title: "4NQO OSCC Bcat - scRNA-seq QC and Preprocessing"
author: "M. Muzamil Khan"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    theme: flatly
    toc: yes
    toc_float: true
  html_notebook:
    code_folding: hide
    theme: flatly
    toc: yes
    toc_float: true
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath("../.."))
```

```{r libraries, warning=FALSE, message=FALSE}
library(Seurat)
library(ggplot2)
library(dplyr)
library(DT)
library(patchwork)
PATH <- getwd()
```

## Overview

This document performs quality control and preprocessing of the single-cell RNA-seq data from the 4NQO OSCC / beta-catenin perturbation study. Raw count matrices are loaded from 10x Genomics Cell Ranger output and processed through the Seurat pipeline.

Samples included:

- **Control**: Wild-type mice, no treatment
- **4NQO**: Wild-type mice, 4NQO-induced OSCC
- **4NQO_CPB**: 4NQO-treated mice with conditional Ctnnb1 (beta-catenin) knockout

```{r sample-metadata}
sample_meta <- data.frame(
  sample_id = c("Control_1", "Control_2",
                 "4NQO_1",    "4NQO_2",
                 "4NQO_CPB_1","4NQO_CPB_2"),
  group     = c("Control",  "Control",
                 "4NQO",     "4NQO",
                 "4NQO_CPB", "4NQO_CPB"),
  path      = file.path(PATH, "data", "cellranger",
                         c("Control_1", "Control_2",
                           "4NQO_1",    "4NQO_2",
                           "4NQO_CPB_1","4NQO_CPB_2"),
                         "filtered_feature_bc_matrix")
)
DT::datatable(sample_meta[, c("sample_id", "group")],
               caption = "Sample Metadata")
```

## Load and Merge Seurat Objects {.tabset}

### Load Individual Samples

```{r load-samples}
seurat_list <- lapply(seq_len(nrow(sample_meta)), function(i) {
  counts <- Read10X(data.dir = sample_meta$path[i])
  so     <- CreateSeuratObject(counts      = counts,
                                project     = sample_meta$sample_id[i],
                                min.cells   = 3,
                                min.features = 200)
  so$sample_id <- sample_meta$sample_id[i]
  so$group     <- sample_meta$group[i]
  so
})
names(seurat_list) <- sample_meta$sample_id
```

### Cell Counts per Sample

```{r cell-counts}
cell_counts <- sapply(seurat_list, ncol)
DT::datatable(data.frame(Sample = names(cell_counts), Cells = cell_counts),
               caption = "Cell counts before QC filtering")
```

## Quality Control {.tabset .tabset-fade .tabset-pills}

### Mitochondrial Percentage

```{r mito-pct}
seurat_list <- lapply(seurat_list, function(so) {
  so[["pct_mt"]] <- PercentageFeatureSet(so, pattern = "^mt-")
  so
})

# Violin plots per sample
lapply(seurat_list, function(so) {
  VlnPlot(so,
           features = c("nFeature_RNA", "nCount_RNA", "pct_mt"),
           ncol     = 3,
           pt.size  = 0) +
    plot_annotation(title = unique(so$sample_id))
})
```

### Feature-Count Scatter

```{r feature-count-scatter}
lapply(seurat_list, function(so) {
  p1 <- FeatureScatter(so, feature1 = "nCount_RNA",   feature2 = "pct_mt")
  p2 <- FeatureScatter(so, feature1 = "nCount_RNA",   feature2 = "nFeature_RNA")
  p1 + p2
})
```

## Filtering {.tabset}

### Apply QC Thresholds

```{r filtering}
min_features <- 200
max_features <- 6000
max_mt       <- 20

seurat_list_filtered <- lapply(seurat_list, function(so) {
  subset(so,
          subset = nFeature_RNA >= min_features &
                   nFeature_RNA <= max_features &
                   pct_mt       <= max_mt)
})

cell_counts_after <- sapply(seurat_list_filtered, ncol)
comparison <- data.frame(
  Sample  = names(cell_counts),
  Before  = cell_counts,
  After   = cell_counts_after,
  Removed = cell_counts - cell_counts_after
)
DT::datatable(comparison, caption = "Cell counts before and after QC filtering")
```

## Normalization and Feature Selection {.tabset}

### Normalize and Find HVGs

```{r normalize}
seurat_list_filtered <- lapply(seurat_list_filtered, function(so) {
  so <- NormalizeData(so, normalization.method = "LogNormalize",
                       scale.factor = 10000, verbose = FALSE)
  so <- FindVariableFeatures(so, selection.method = "vst",
                              nfeatures = 3000,    verbose = FALSE)
  so
})
```

### Top HVGs

```{r hvg-plot}
top_hvg <- head(VariableFeatures(seurat_list_filtered[[1]]), 20)
p <- VariableFeaturePlot(seurat_list_filtered[[1]])
LabelPoints(plot = p, points = top_hvg, repel = TRUE) +
  labs(title = paste("Top HVGs -", names(seurat_list_filtered)[1]))
```

## Merge and Scale {.tabset}

### Merge Seurat Objects

```{r merge}
seurat_merged <- merge(seurat_list_filtered[[1]],
                        y           = seurat_list_filtered[-1],
                        add.cell.ids = sample_meta$sample_id,
                        project     = "4NQO_OSCC_Bcat",
                        merge.data  = TRUE)

seurat_merged$group <- factor(seurat_merged$group,
                               levels = c("Control", "4NQO", "4NQO_CPB"))
print(seurat_merged)
```

### Scale Data

```{r scale}
all_genes       <- rownames(seurat_merged)
seurat_merged   <- ScaleData(seurat_merged, features = all_genes, verbose = FALSE)
```

## Dimensionality Reduction {.tabset .tabset-fade .tabset-pills}

### PCA

```{r pca}
seurat_merged <- RunPCA(seurat_merged, features = VariableFeatures(seurat_merged),
                         npcs = 50, verbose = FALSE)
ElbowPlot(seurat_merged, ndims = 50) +
  labs(title = "Elbow Plot - PCA")
```

### PCA Visualization

```{r pca-viz}
DimPlot(seurat_merged, reduction = "pca", group.by = "group") +
  labs(title = "PCA - Colored by Group")
DimPlot(seurat_merged, reduction = "pca", group.by = "sample_id") +
  labs(title = "PCA - Colored by Sample")
```

### UMAP

```{r umap}
n_pcs           <- 30
seurat_merged   <- RunUMAP(seurat_merged, dims = 1:n_pcs, verbose = FALSE)
DimPlot(seurat_merged, reduction = "umap", group.by = "group") +
  labs(title = "UMAP - Colored by Group")
DimPlot(seurat_merged, reduction = "umap", group.by = "sample_id") +
  labs(title = "UMAP - Colored by Sample")
```

## Save Preprocessed Object {.tabset}

```{r save, eval=FALSE}
saveRDS(seurat_merged, file.path(PATH, "data/seurat_preprocessed.RDS"))
message("Saved preprocessed Seurat object: ",
         ncol(seurat_merged), " cells x ", nrow(seurat_merged), " genes")
```

## Session Info

```{r session-info}
sessionInfo()
```
