---
title: "4NQO OSCC Bcat - scRNA-seq Clustering and Cell Type Annotation"
author: "M. Muzamil Khan"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    theme: flatly
    toc: yes
    toc_float: true
  html_notebook:
    code_folding: hide
    theme: flatly
    toc: yes
    toc_float: true
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath("../.."))
```

```{r libraries, warning=FALSE, message=FALSE}
library(Seurat)
library(ggplot2)
library(dplyr)
library(DT)
library(patchwork)
library(RColorBrewer)
PATH <- getwd()
```

## Overview

This document performs unsupervised clustering and cell type annotation of the preprocessed single-cell RNA-seq data. QC and preprocessing steps are described in `01_qc_preprocessing.{Rmd, html}`.

Cell type annotation is performed using a combination of:

1. **Canonical marker genes** for oral tissue cell types (epithelial, fibroblast, immune, endothelial)
2. **Cluster marker genes** identified by Seurat's `FindAllMarkers`

```{r load-data}
seurat_merged <- readRDS(file.path(PATH, "data/seurat_preprocessed.RDS"))
print(seurat_merged)
table(seurat_merged$group)
```

## Clustering {.tabset .tabset-fade .tabset-pills}

### Find Neighbors and Clusters

```{r clustering}
n_pcs      <- 30
resolution <- 0.5

seurat_merged <- FindNeighbors(seurat_merged, dims = 1:n_pcs, verbose = FALSE)
seurat_merged <- FindClusters(seurat_merged,  resolution = resolution, verbose = FALSE)

cat("Number of clusters:", length(unique(Idents(seurat_merged))), "\n")
table(Idents(seurat_merged))
```

### UMAP by Cluster

```{r umap-cluster}
DimPlot(seurat_merged, reduction = "umap", label = TRUE, repel = TRUE) +
  labs(title = paste0("UMAP - Clusters (resolution = ", resolution, ")")) +
  NoLegend()
```

### UMAP by Group

```{r umap-group}
DimPlot(seurat_merged, reduction = "umap", group.by = "group") +
  labs(title = "UMAP - Colored by Experimental Group")
```

### UMAP by Sample

```{r umap-sample}
DimPlot(seurat_merged, reduction = "umap", group.by = "sample_id") +
  labs(title = "UMAP - Colored by Sample")
```

### Cluster Composition

```{r cluster-composition}
comp <- table(seurat_merged$seurat_clusters, seurat_merged$group)
comp_pct <- prop.table(comp, margin = 1) * 100
comp_df  <- as.data.frame(comp_pct)
names(comp_df) <- c("Cluster", "Group", "Pct")

ggplot(comp_df, aes(x = Cluster, y = Pct, fill = Group)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Cluster Composition by Group",
       y = "Percentage (%)", x = "Cluster") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Cluster Marker Genes {.tabset .tabset-fade .tabset-pills}

### FindAllMarkers

```{r find-markers}
all_markers <- FindAllMarkers(seurat_merged,
                               only.pos     = TRUE,
                               min.pct      = 0.25,
                               logfc.threshold = 0.25,
                               verbose      = FALSE)

top_markers <- all_markers %>%
  group_by(cluster) %>%
  slice_max(n = 10, order_by = avg_log2FC)

DT::datatable(top_markers, caption = "Top 10 markers per cluster (by avg_log2FC)")

saveRDS(all_markers, file.path(PATH, "results/sc_cluster_markers.RDS"))
```

### Dot Plot - Top Markers

```{r dotplot-markers}
top5 <- all_markers %>%
  group_by(cluster) %>%
  slice_max(n = 5, order_by = avg_log2FC) %>%
  pull(gene) %>%
  unique()

DotPlot(seurat_merged, features = top5) +
  RotatedAxis() +
  labs(title = "Top 5 Markers per Cluster")
```

## Cell Type Annotation {.tabset .tabset-fade .tabset-pills}

### Canonical Markers

```{r canonical-markers}
canonical_markers <- list(
  Epithelial    = c("Epcam", "Krt5", "Krt14", "Krt8", "Krt18", "Cdh1"),
  Fibroblast    = c("Col1a1", "Col1a2", "Vim", "Pdgfra", "Fap"),
  Endothelial   = c("Pecam1", "Cdh5", "Vwf", "Eng"),
  T_cell        = c("Cd3e", "Cd3d", "Cd4", "Cd8a", "Foxp3"),
  B_cell        = c("Cd19", "Ms4a1", "Cd79a"),
  Myeloid       = c("Cd68", "Csf1r", "Lyz2", "Adgre1"),
  NK_cell       = c("Nkg7", "Klrd1", "Gzmb"),
  Mast_cell     = c("Mcpt8", "Cpa3", "Kit")
)

FeaturePlot(seurat_merged,
             features  = c("Epcam", "Col1a1", "Pecam1", "Cd3e",
                            "Cd19",  "Cd68",   "Nkg7",   "Cpa3"),
             ncol      = 4,
             reduction = "umap")
```

### Marker Heatmap

```{r marker-heatmap}
DoHeatmap(seurat_merged,
           features = unlist(canonical_markers),
           group.by = "seurat_clusters",
           slot     = "scale.data") +
  theme(axis.text.y = element_text(size = 6))
```

### Assign Cell Types

```{r assign-celltypes}
# Assign cell type labels based on marker expression and cluster inspection
# (To be updated after visual inspection of marker plots)
cluster_to_celltype <- c(
  "0"  = "Epithelial",
  "1"  = "Epithelial",
  "2"  = "Fibroblast",
  "3"  = "T_cell",
  "4"  = "Myeloid",
  "5"  = "Epithelial",
  "6"  = "Endothelial",
  "7"  = "B_cell",
  "8"  = "NK_cell",
  "9"  = "Mast_cell",
  "10" = "Fibroblast"
)

seurat_merged$cell_type <- cluster_to_celltype[as.character(seurat_merged$seurat_clusters)]
seurat_merged$cell_type <- factor(seurat_merged$cell_type,
                                   levels = c("Epithelial", "Fibroblast", "Endothelial",
                                              "T_cell", "B_cell", "Myeloid",
                                              "NK_cell", "Mast_cell"))
```

### UMAP by Cell Type

```{r umap-celltype}
cell_type_colors <- c(
  Epithelial   = "#E41A1C",
  Fibroblast   = "#FF7F00",
  Endothelial  = "#984EA3",
  T_cell       = "#4DAF4A",
  B_cell       = "#377EB8",
  Myeloid      = "#A65628",
  NK_cell      = "#F781BF",
  Mast_cell    = "#999999"
)

DimPlot(seurat_merged, reduction = "umap", group.by = "cell_type",
         cols = cell_type_colors, label = TRUE, repel = TRUE) +
  labs(title = "UMAP - Cell Type Annotation") +
  NoLegend()
```

### Cell Type Composition by Group

```{r celltype-composition}
ct_comp <- table(seurat_merged$cell_type, seurat_merged$group)
ct_pct  <- prop.table(ct_comp, margin = 2) * 100
ct_df   <- as.data.frame(ct_pct)
names(ct_df) <- c("CellType", "Group", "Pct")

ggplot(ct_df, aes(x = Group, y = Pct, fill = CellType)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = cell_type_colors) +
  labs(title = "Cell Type Composition by Experimental Group",
       y = "Percentage (%)", x = "Group") +
  theme_bw()
```

## Beta-Catenin Pathway Activity {.tabset}

### Wnt Target Gene Expression

```{r wnt-markers}
wnt_targets <- c("Axin2", "Lgr5", "Myc", "Ccnd1", "Cd44", "Tcf7", "Lef1", "Nkd1")
wnt_targets_present <- intersect(wnt_targets, rownames(seurat_merged))

FeaturePlot(seurat_merged, features = wnt_targets_present, ncol = 4, reduction = "umap")
```

### Ctnnb1 Expression

```{r ctnnb1-expression}
FeaturePlot(seurat_merged, features = "Ctnnb1", reduction = "umap") +
  labs(title = "Ctnnb1 (Beta-catenin) Expression")

VlnPlot(seurat_merged, features = "Ctnnb1", group.by = "group", pt.size = 0) +
  labs(title = "Ctnnb1 Expression by Group")
```

## Save Annotated Object {.tabset}

```{r save, eval=FALSE}
saveRDS(seurat_merged, file.path(PATH, "data/seurat_annotated.RDS"))
message("Saved annotated Seurat object")
```

## Session Info

```{r session-info}
sessionInfo()
```
