---
title: "4NQO OSCC Bcat - scRNA-seq Differential Expression Analysis"
author: "M. Muzamil Khan"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    theme: flatly
    toc: yes
    toc_float: true
  html_notebook:
    code_folding: hide
    theme: flatly
    toc: yes
    toc_float: true
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath("../.."))
```

```{r libraries, warning=FALSE, message=FALSE}
library(Seurat)
library(ggplot2)
library(dplyr)
library(DT)
library(patchwork)
library(MAST)
PATH <- getwd()
```

## Overview

This document performs differential expression analysis between experimental groups at the single-cell level using MAST, applied within each major cell type. Cell type annotations are described in `02_clustering.{Rmd, html}`.

Comparisons performed per cell type:

- **4NQO vs. Control**
- **4NQO + CPB vs. 4NQO**
- **4NQO + CPB vs. Control**

```{r load-data}
seurat_merged <- readRDS(file.path(PATH, "data/seurat_annotated.RDS"))
print(seurat_merged)
table(seurat_merged$group, seurat_merged$cell_type)
```

```{r helper-functions}
run_mast <- function(seurat_obj, cell_type_col = "cell_type",
                      group_col = "group", ident1, ident2,
                      cell_type = NULL, min_cells = 10) {
  if (!is.null(cell_type)) {
    seurat_obj <- subset(seurat_obj, cells = which(seurat_obj[[cell_type_col]][, 1] == cell_type))
  }
  cells_g1 <- colnames(seurat_obj)[seurat_obj[[group_col]][, 1] == ident1]
  cells_g2 <- colnames(seurat_obj)[seurat_obj[[group_col]][, 1] == ident2]
  if (length(cells_g1) < min_cells || length(cells_g2) < min_cells) {
    message("Skipping: not enough cells for ", cell_type, " (", ident1, " vs. ", ident2, ")")
    return(NULL)
  }
  sub_obj <- subset(seurat_obj, cells = c(cells_g1, cells_g2))
  Idents(sub_obj) <- sub_obj[[group_col]][, 1]
  FindMarkers(sub_obj,
               ident.1 = ident1, ident.2 = ident2,
               test.use = "MAST",
               min.pct  = 0.1,
               verbose  = FALSE)
}

cell_types <- levels(seurat_merged$cell_type)
```

## 4NQO vs. Control - Per Cell Type {.tabset .tabset-fade .tabset-pills}

```{r de-oscc-ctrl, results='asis'}
de_oscc_ctrl <- list()

for (ct in cell_types) {
  cat("### ", ct, "\n\n")
  res <- run_mast(seurat_merged,
                   ident1    = "4NQO",
                   ident2    = "Control",
                   cell_type = ct)
  if (!is.null(res) && nrow(res) > 0) {
    res$gene      <- rownames(res)
    res$cell_type <- ct
    de_oscc_ctrl[[ct]] <- res
    sig <- res[res$p_val_adj <= 0.05 & abs(res$avg_log2FC) >= 0.5, ]
    cat("Significant DEGs (padj <= 0.05, |log2FC| >= 0.5):", nrow(sig), "\n\n")
    if (nrow(sig) > 0) {
      print(DT::datatable(sig, caption = paste(ct, "- 4NQO vs. Control")))
    }
  } else {
    cat("No significant DEGs or insufficient cells.\n\n")
  }
  cat("\n\n")
}

de_oscc_ctrl_all <- do.call(rbind, de_oscc_ctrl)
saveRDS(de_oscc_ctrl_all, file.path(PATH, "results/sc_de_4nqo_vs_ctrl.RDS"))
```

## 4NQO + CPB vs. 4NQO - Per Cell Type {.tabset .tabset-fade .tabset-pills}

```{r de-cpb-oscc, results='asis'}
de_cpb_oscc <- list()

for (ct in cell_types) {
  cat("### ", ct, "\n\n")
  res <- run_mast(seurat_merged,
                   ident1    = "4NQO_CPB",
                   ident2    = "4NQO",
                   cell_type = ct)
  if (!is.null(res) && nrow(res) > 0) {
    res$gene      <- rownames(res)
    res$cell_type <- ct
    de_cpb_oscc[[ct]] <- res
    sig <- res[res$p_val_adj <= 0.05 & abs(res$avg_log2FC) >= 0.5, ]
    cat("Significant DEGs (padj <= 0.05, |log2FC| >= 0.5):", nrow(sig), "\n\n")
    if (nrow(sig) > 0) {
      print(DT::datatable(sig, caption = paste(ct, "- 4NQO + CPB vs. 4NQO")))
    }
  } else {
    cat("No significant DEGs or insufficient cells.\n\n")
  }
  cat("\n\n")
}

de_cpb_oscc_all <- do.call(rbind, de_cpb_oscc)
saveRDS(de_cpb_oscc_all, file.path(PATH, "results/sc_de_4nqo_cpb_vs_4nqo.RDS"))
```

## 4NQO + CPB vs. Control - Per Cell Type {.tabset .tabset-fade .tabset-pills}

```{r de-cpb-ctrl, results='asis'}
de_cpb_ctrl <- list()

for (ct in cell_types) {
  cat("### ", ct, "\n\n")
  res <- run_mast(seurat_merged,
                   ident1    = "4NQO_CPB",
                   ident2    = "Control",
                   cell_type = ct)
  if (!is.null(res) && nrow(res) > 0) {
    res$gene      <- rownames(res)
    res$cell_type <- ct
    de_cpb_ctrl[[ct]] <- res
    sig <- res[res$p_val_adj <= 0.05 & abs(res$avg_log2FC) >= 0.5, ]
    cat("Significant DEGs (padj <= 0.05, |log2FC| >= 0.5):", nrow(sig), "\n\n")
    if (nrow(sig) > 0) {
      print(DT::datatable(sig, caption = paste(ct, "- 4NQO + CPB vs. Control")))
    }
  } else {
    cat("No significant DEGs or insufficient cells.\n\n")
  }
  cat("\n\n")
}

de_cpb_ctrl_all <- do.call(rbind, de_cpb_ctrl)
saveRDS(de_cpb_ctrl_all, file.path(PATH, "results/sc_de_4nqo_cpb_vs_ctrl.RDS"))
```

## Visualization {.tabset .tabset-fade .tabset-pills}

### Top DEGs - Epithelial Cells (4NQO vs. Control)

```{r epi-violin}
if (!is.null(de_oscc_ctrl[["Epithelial"]])) {
  top_epi_genes <- de_oscc_ctrl[["Epithelial"]] %>%
    filter(p_val_adj <= 0.05, avg_log2FC >= 0.5) %>%
    slice_max(n = 6, order_by = avg_log2FC) %>%
    pull(gene)

  epi_cells <- subset(seurat_merged, subset = cell_type == "Epithelial")
  VlnPlot(epi_cells, features = top_epi_genes,
           group.by = "group", ncol = 3, pt.size = 0) +
    plot_annotation(title = "Top Upregulated Genes - Epithelial Cells (4NQO vs. Control)")
}
```

### Wnt Targets in Epithelial Cells

```{r wnt-epi}
wnt_targets <- c("Axin2", "Lgr5", "Myc", "Ccnd1", "Ctnnb1", "Cd44")
epi_cells   <- subset(seurat_merged, subset = cell_type == "Epithelial")

VlnPlot(epi_cells, features = intersect(wnt_targets, rownames(epi_cells)),
         group.by = "group", ncol = 3, pt.size = 0) +
  plot_annotation(title = "Wnt Target Gene Expression in Epithelial Cells")
```

### DEG Summary Across Cell Types

```{r deg-summary}
summarize_de <- function(de_list, comparison) {
  do.call(rbind, lapply(names(de_list), function(ct) {
    res <- de_list[[ct]]
    sig <- res[res$p_val_adj <= 0.05 & abs(res$avg_log2FC) >= 0.5, ]
    data.frame(
      Comparison = comparison,
      CellType   = ct,
      N_up       = sum(sig$avg_log2FC > 0),
      N_down     = sum(sig$avg_log2FC < 0),
      Total      = nrow(sig)
    )
  }))
}

deg_summary <- rbind(
  summarize_de(de_oscc_ctrl, "4NQO vs. Control"),
  summarize_de(de_cpb_oscc,  "4NQO_CPB vs. 4NQO"),
  summarize_de(de_cpb_ctrl,  "4NQO_CPB vs. Control")
)

DT::datatable(deg_summary, caption = "DEG Summary Across Cell Types and Comparisons")

ggplot(deg_summary, aes(x = CellType, y = Total, fill = Comparison)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Number of Significant DEGs per Cell Type",
       x = "Cell Type", y = "# DEGs (padj <= 0.05, |log2FC| >= 0.5)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Session Info

```{r session-info}
sessionInfo()
```
