---
title: "4NQO OSCC Bcat - scRNA-seq Pathway Enrichment Analysis"
author: "M. Muzamil Khan"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    theme: flatly
    toc: yes
    toc_float: true
  html_notebook:
    code_folding: hide
    theme: flatly
    toc: yes
    toc_float: true
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath("../.."))
```

```{r libraries, warning=FALSE, message=FALSE}
library(Seurat)
library(ggplot2)
library(dplyr)
library(DT)
library(hypeR)
library(patchwork)
PATH <- getwd()
```

## Overview

This document performs pathway enrichment analysis on single-cell differential expression results using hypeR (ORA) and gene module scoring. Single-cell DE results are loaded from `03_diffanalysis.{Rmd, html}`.

Analyses performed:

1. **ORA** on cell-type-specific DEGs using MSigDB Hallmarks and KEGG pathways
2. **Module scoring** of Wnt/beta-catenin and cancer-related gene sets across cells
3. **Cell proportion analysis** (differential abundance) between experimental groups

```{r load-data}
seurat_merged <- readRDS(file.path(PATH, "data/seurat_annotated.RDS"))
de_oscc_ctrl  <- readRDS(file.path(PATH, "results/sc_de_4nqo_vs_ctrl.RDS"))
de_cpb_oscc   <- readRDS(file.path(PATH, "results/sc_de_4nqo_cpb_vs_4nqo.RDS"))
de_cpb_ctrl   <- readRDS(file.path(PATH, "results/sc_de_4nqo_cpb_vs_ctrl.RDS"))

cell_types <- levels(seurat_merged$cell_type)
```

```{r genesets}
msigdb_H  <- msigdb_gsets(species = "Mus musculus", category = "H")
msigdb_C2 <- msigdb_gsets(species = "Mus musculus", category = "C2",
                            subcategory = "CP:KEGG")
```

## ORA: 4NQO vs. Control - Epithelial Cells {.tabset .tabset-fade .tabset-pills}

```{r ora-epi-oscc-ctrl}
epi_up   <- de_oscc_ctrl %>%
  filter(cell_type == "Epithelial", p_val_adj <= 0.05, avg_log2FC >= 0.5) %>%
  pull(gene)
epi_down <- de_oscc_ctrl %>%
  filter(cell_type == "Epithelial", p_val_adj <= 0.05, avg_log2FC <= -0.5) %>%
  pull(gene)
```

### Hallmarks - Upregulated

```{r ora-epi-up-H}
if (length(epi_up) > 0) {
  hyp_epi_up_H <- hypeR(epi_up,
                          gsets      = msigdb_H,
                          test       = "hypergeometric",
                          background = nrow(seurat_merged),
                          fdr        = 0.05,
                          quiet      = TRUE)
  hyp_dots(hyp_epi_up_H, title = "Epithelial (Up) - 4NQO vs. Control - Hallmarks")
} else {
  cat("No upregulated genes to test.\n")
}
```

### Hallmarks - Downregulated

```{r ora-epi-down-H}
if (length(epi_down) > 0) {
  hyp_epi_down_H <- hypeR(epi_down,
                            gsets      = msigdb_H,
                            test       = "hypergeometric",
                            background = nrow(seurat_merged),
                            fdr        = 0.05,
                            quiet      = TRUE)
  hyp_dots(hyp_epi_down_H, title = "Epithelial (Down) - 4NQO vs. Control - Hallmarks")
} else {
  cat("No downregulated genes to test.\n")
}
```

### KEGG - Upregulated

```{r ora-epi-up-C2}
if (length(epi_up) > 0) {
  hyp_epi_up_C2 <- hypeR(epi_up,
                           gsets      = msigdb_C2,
                           test       = "hypergeometric",
                           background = nrow(seurat_merged),
                           fdr        = 0.05,
                           quiet      = TRUE)
  hyp_dots(hyp_epi_up_C2, title = "Epithelial (Up) - 4NQO vs. Control - KEGG")
}
```

## ORA: 4NQO + CPB vs. 4NQO - Epithelial Cells {.tabset .tabset-fade .tabset-pills}

```{r ora-cpb-epi}
cpb_epi_up   <- de_cpb_oscc %>%
  filter(cell_type == "Epithelial", p_val_adj <= 0.05, avg_log2FC >= 0.5) %>%
  pull(gene)
cpb_epi_down <- de_cpb_oscc %>%
  filter(cell_type == "Epithelial", p_val_adj <= 0.05, avg_log2FC <= -0.5) %>%
  pull(gene)
```

### Hallmarks - Upregulated

```{r ora-cpb-epi-up-H}
if (length(cpb_epi_up) > 0) {
  hyp_cpb_epi_up_H <- hypeR(cpb_epi_up,
                              gsets      = msigdb_H,
                              test       = "hypergeometric",
                              background = nrow(seurat_merged),
                              fdr        = 0.05,
                              quiet      = TRUE)
  hyp_dots(hyp_cpb_epi_up_H, title = "Epithelial (Up) - 4NQO_CPB vs. 4NQO - Hallmarks")
} else {
  cat("No upregulated genes to test.\n")
}
```

### Hallmarks - Downregulated

```{r ora-cpb-epi-down-H}
if (length(cpb_epi_down) > 0) {
  hyp_cpb_epi_down_H <- hypeR(cpb_epi_down,
                                gsets      = msigdb_H,
                                test       = "hypergeometric",
                                background = nrow(seurat_merged),
                                fdr        = 0.05,
                                quiet      = TRUE)
  hyp_dots(hyp_cpb_epi_down_H, title = "Epithelial (Down) - 4NQO_CPB vs. 4NQO - Hallmarks")
} else {
  cat("No downregulated genes to test.\n")
}
```

## Gene Module Scoring {.tabset .tabset-fade .tabset-pills}

### Wnt / Beta-catenin Module Score

```{r wnt-module-score}
wnt_module <- list(
  Wnt_targets = c("Axin2", "Lgr5", "Myc", "Ccnd1", "Cd44",
                   "Tcf7",  "Lef1", "Nkd1", "Sp5",   "Dkk1")
)
seurat_merged <- AddModuleScore(seurat_merged,
                                 features = wnt_module,
                                 name     = "Wnt_score",
                                 nbin     = 24,
                                 ctrl     = 100)

FeaturePlot(seurat_merged, features = "Wnt_score1", reduction = "umap",
             min.cutoff = "q5", max.cutoff = "q95") +
  labs(title = "Wnt/Beta-catenin Module Score")

VlnPlot(seurat_merged, features = "Wnt_score1",
         group.by = "group", pt.size = 0) +
  labs(title = "Wnt Module Score by Group")
```

### EMT Module Score

```{r emt-module-score}
emt_module <- list(
  EMT = c("Cdh2", "Vim", "Fn1", "Zeb1", "Zeb2",
           "Twist1", "Snai1", "Snai2", "Mmp2", "Mmp9")
)
seurat_merged <- AddModuleScore(seurat_merged,
                                 features = emt_module,
                                 name     = "EMT_score",
                                 nbin     = 24,
                                 ctrl     = 100)

FeaturePlot(seurat_merged, features = "EMT_score1", reduction = "umap",
             min.cutoff = "q5", max.cutoff = "q95") +
  labs(title = "EMT Module Score")

VlnPlot(seurat_merged, features = "EMT_score1",
         group.by = "group", pt.size = 0) +
  labs(title = "EMT Module Score by Group")
```

### Proliferation Module Score

```{r prolif-module-score}
prolif_module <- list(
  Proliferation = c("Mki67", "Top2a", "Pcna", "Cdk1", "Ccnb1",
                     "Ccna2", "Birc5", "Ube2c", "Cenpf", "Plk1")
)
seurat_merged <- AddModuleScore(seurat_merged,
                                 features = prolif_module,
                                 name     = "Prolif_score",
                                 nbin     = 24,
                                 ctrl     = 100)

FeaturePlot(seurat_merged, features = "Prolif_score1", reduction = "umap",
             min.cutoff = "q5", max.cutoff = "q95") +
  labs(title = "Proliferation Module Score")
```

### Wnt Score by Cell Type and Group

```{r wnt-score-celltype}
ggplot(seurat_merged@meta.data,
        aes(x = group, y = Wnt_score1, fill = group)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  facet_wrap(~cell_type, scales = "free_y") +
  labs(title = "Wnt Module Score by Cell Type and Group",
       x = "Group", y = "Wnt Score") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

## Cell Proportion Analysis {.tabset}

### Proportion Changes

```{r cell-proportions}
prop_table <- seurat_merged@meta.data %>%
  group_by(group, cell_type) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(group) %>%
  mutate(pct = n / sum(n) * 100)

ggplot(prop_table, aes(x = group, y = pct, fill = cell_type)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Cell Type Proportions by Group",
       x = "Group", y = "Proportion (%)") +
  theme_bw()

DT::datatable(prop_table, caption = "Cell Type Proportions by Experimental Group")
```

## Save Final Object

```{r save, eval=FALSE}
saveRDS(seurat_merged, file.path(PATH, "data/seurat_final.RDS"))
message("Saved final Seurat object with module scores")
```

## Session Info

```{r session-info}
sessionInfo()
```
