---
title: "4NQO OSCC Bcat - Bulk RNA-seq Summary Statistics and eSet Creation"
author: "M. Muzamil Khan"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    theme: flatly
    toc: yes
    toc_float: true
  html_notebook:
    code_folding: hide
    theme: flatly
    toc: yes
    toc_float: true
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath("../.."))
```

```{r libraries, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(DT)
library(Biobase)
library(reshape2)
library(hypeR)
PATH <- getwd()
```

## Overview

This document summarizes the bulk RNA-seq data quality metrics and creates the ExpressionSet object used in downstream analyses for the 4NQO-induced oral squamous cell carcinoma (OSCC) study with conditional beta-catenin perturbation.

The experiment includes the following groups:

- **Control**: Wild-type mice without 4NQO treatment
- **4NQO**: Wild-type mice treated with 4NQO to induce OSCC
- **4NQO + CPB**: 4NQO-treated mice with conditional perturbation of beta-catenin (Ctnnb1)

Paired-end reads were aligned to the mouse genome (mm10/GRCm38) using STAR and quantified with featureCounts.

## Load Data {.tabset}

### Feature Counts Metrics

```{r load-data}
featureCounts_metrics <- read.table(file.path(PATH, "data/featureCounts_metrics.txt"),
                                     header = TRUE, sep = "\t")
annotData <- read.csv(file.path(PATH, "data/sample_annotations.csv"), header = TRUE)
assaydata  <- read.table(file.path(PATH, "data/features_counts.txt"),
                          header = TRUE, row.names = 1)

rownames(annotData) <- annotData$Sample_ID
colnames(assaydata) <- gsub("\\.", "_", colnames(assaydata))

annotData$Group <- factor(annotData$Group, levels = c("Control", "4NQO", "4NQO_CPB"))
```

### Sample Annotations

```{r annotations}
DT::datatable(annotData, caption = "Sample Annotations")
```

## Reads Summary {.tabset .tabset-fade .tabset-pills}

### Reads per Sample

```{r reads-per-sample}
fc_trans <- data.frame(t(featureCounts_metrics))
rownames(fc_trans) <- gsub("\\.", "_", rownames(fc_trans))
fc_trans <- data.frame(fc_trans,
                        annotData[match(rownames(fc_trans), rownames(annotData)), ])
fc_trans$mapping_rate <- (fc_trans$Mapped_fragments / fc_trans$Total_fragments) * 100
fc_trans$rawcounts    <- colSums(assaydata)[match(colnames(assaydata), rownames(fc_trans))]

ggplot(data = fc_trans,
       aes(x = reorder(Sample_ID, -Total_fragments), y = Total_fragments, fill = Group)) +
  geom_bar(stat = "identity", width = 0.75) +
  labs(title = "Total Reads per Sample", y = "Reads", x = "Sample ID") +
  coord_flip() +
  theme_bw() +
  theme(axis.text = element_text(size = 7),
        axis.title = element_text(size = 10, face = "bold"))
```

### Mapping Rate

```{r mapping-rate}
ggplot(fc_trans,
       aes(reorder(Group, mapping_rate, median, order = TRUE), mapping_rate, fill = Group)) +
  geom_boxplot() +
  labs(title = "Mapping Rate by Group", y = "Mapping Rate (%)", x = "Group") +
  theme_bw()
```

### Raw Counts

```{r raw-counts}
ggplot(fc_trans,
       aes(reorder(Group, rawcounts, median, order = TRUE), rawcounts, fill = Group)) +
  geom_boxplot() +
  labs(title = "Raw Counts by Group", y = "Raw Counts", x = "Group") +
  theme_bw()
```

## Quality Control {.tabset .tabset-fade .tabset-pills}

### Library Size Distribution

```{r lib-size}
lib_sizes <- data.frame(
  Sample_ID = colnames(assaydata),
  LibSize   = colSums(assaydata)
)
lib_sizes <- merge(lib_sizes, annotData, by = "Sample_ID")

ggplot(lib_sizes, aes(x = reorder(Sample_ID, -LibSize), y = LibSize, fill = Group)) +
  geom_bar(stat = "identity", width = 0.75) +
  labs(title = "Library Size per Sample", y = "Total Counts", x = "Sample ID") +
  coord_flip() +
  theme_bw() +
  theme(axis.text = element_text(size = 7))
```

### Detected Genes

```{r detected-genes}
detected_genes <- data.frame(
  Sample_ID      = colnames(assaydata),
  Detected_genes = colSums(assaydata > 0)
)
detected_genes <- merge(detected_genes, annotData, by = "Sample_ID")

ggplot(detected_genes,
       aes(x = reorder(Group, Detected_genes, median, order = TRUE),
           y = Detected_genes, fill = Group)) +
  geom_boxplot() +
  labs(title = "Detected Genes by Group", y = "Number of Detected Genes", x = "Group") +
  theme_bw()
```

## ExpressionSet Creation {.tabset}

### Build eSet

```{r eset-creation, eval=FALSE}
eSet <- Biobase::ExpressionSet(
  assayData  = as.matrix(assaydata),
  phenoData  = AnnotatedDataFrame(
    annotData[match(colnames(assaydata), rownames(annotData)), ]
  )
)

saveRDS(eSet, file.path(PATH, "data/eSet_bulk.RDS"))
message("eSet saved: ", nrow(eSet), " genes x ", ncol(eSet), " samples")
```

### eSet Summary

```{r eset-summary, eval=FALSE}
eSet <- readRDS(file.path(PATH, "data/eSet_bulk.RDS"))
print(eSet)
table(eSet$Group)
```

## Session Info

```{r session-info}
sessionInfo()
```
