---
title: "4NQO OSCC Bcat - Bulk RNA-seq Differential Expression Analysis"
author: "M. Muzamil Khan"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    theme: flatly
    toc: yes
    toc_float: true
  html_notebook:
    code_folding: hide
    theme: flatly
    toc: yes
    toc_float: true
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath("../.."))
```

```{r libraries, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(DT)
library(Biobase)
library(DESeq2)
library(pheatmap)
library(RColorBrewer)
library(hypeR)
PATH <- getwd()
```

## Overview

This document performs pairwise differential expression analysis using DESeq2 between the experimental groups in the 4NQO OSCC/beta-catenin study. Summary statistics and eSet creation are in `01_summstats_eset.{Rmd, html}`.

Groups compared:

- **4NQO vs. Control**: Effect of 4NQO-induced carcinogenesis
- **4NQO + CPB vs. 4NQO**: Effect of conditional beta-catenin perturbation in the context of 4NQO-induced OSCC
- **4NQO + CPB vs. Control**: Combined effect of carcinogenesis and Bcat perturbation

```{r load-eset}
eSet <- readRDS(file.path(PATH, "data/eSet_bulk.RDS"))
table(eSet$Group)
```

```{r helper-functions}
# Hierarchical clustering with optimal leaf ordering (S. Monti / D. Gusenleitner)
hcopt <- function(d, HC = NULL, method = "ward.D", members = NULL) {
  require("cba")
  if (is.null(HC)) {
    HC <- hclust(d, method = method, members = members)
  }
  ORD    <- order.optimal(d, merge = HC$merge)
  HC$merge <- ORD$merge
  HC$order <- ORD$order
  HC
}

run_deseq2 <- function(eset, groups, ref_group, covariates = NULL) {
  sub_eset <- eset[, which(pData(eset)$Group %in% groups)]
  col_data <- data.frame(condition = sub_eset$Group, row.names = colnames(sub_eset))
  if (!is.null(covariates)) {
    for (cov in covariates) {
      col_data[[cov]] <- pData(sub_eset)[[cov]]
    }
    design_formula <- as.formula(paste("~", paste(c(covariates, "condition"), collapse = " + ")))
  } else {
    design_formula <- ~ condition
  }
  dds <- DESeqDataSetFromMatrix(countData = exprs(sub_eset),
                                 colData   = col_data,
                                 design    = design_formula)
  dds$condition <- relevel(dds$condition, ref = ref_group)
  dds_res       <- DESeq(dds)
  results(dds_res)
}

plot_volcano <- function(res, title, fdr = 0.05, lfc = 1.5) {
  df <- as.data.frame(res)
  df$sig <- ifelse(!is.na(df$padj) & df$padj <= fdr & abs(df$log2FoldChange) >= lfc,
                   "Significant", "Not significant")
  ggplot(df, aes(x = log2FoldChange, y = -log10(pvalue), color = sig)) +
    geom_point(alpha = 0.5, size = 1) +
    scale_color_manual(values = c("Significant" = "red", "Not significant" = "grey50")) +
    geom_vline(xintercept = c(-lfc, lfc), linetype = "dashed", color = "blue") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "blue") +
    labs(title = title, x = "log2 Fold Change", y = "-log10(p-value)", color = "") +
    theme_bw()
}

fdr   <- 0.05
lfc   <- 1.5
```

## 4NQO vs. Control {.tabset .tabset-fade .tabset-pills}

### DESeq2 Results

```{r oscc-vs-ctrl-deseq2}
res_oscc_ctrl <- run_deseq2(eSet,
                              groups    = c("Control", "4NQO"),
                              ref_group = "Control")
res_oscc_ctrl <- res_oscc_ctrl[!is.na(res_oscc_ctrl$padj), ]
summary(res_oscc_ctrl)

cat("\nSignificant DEGs (padj <= 0.05, |LFC| >= 1.5):\n")
print(c(
  "Up"   = sum(res_oscc_ctrl$padj <= fdr & res_oscc_ctrl$log2FoldChange >= lfc,  na.rm = TRUE),
  "Down" = sum(res_oscc_ctrl$padj <= fdr & res_oscc_ctrl$log2FoldChange <= -lfc, na.rm = TRUE)
))
```

### Volcano Plot

```{r oscc-vs-ctrl-volcano}
plot_volcano(res_oscc_ctrl, title = "4NQO vs. Control")
```

### Top DEGs

```{r oscc-vs-ctrl-top}
oscc_ctrl_up   <- rownames(res_oscc_ctrl[which(res_oscc_ctrl$padj <= fdr &
                                                  res_oscc_ctrl$log2FoldChange >= lfc), ])
oscc_ctrl_down <- rownames(res_oscc_ctrl[which(res_oscc_ctrl$padj <= fdr &
                                                  res_oscc_ctrl$log2FoldChange <= -lfc), ])

DT::datatable(as.data.frame(res_oscc_ctrl[oscc_ctrl_up, ]),
               caption = "4NQO vs. Control - Upregulated")
DT::datatable(as.data.frame(res_oscc_ctrl[oscc_ctrl_down, ]),
               caption = "4NQO vs. Control - Downregulated")
```

### Heatmap

```{r oscc-vs-ctrl-heatmap}
top_genes <- c(oscc_ctrl_up, oscc_ctrl_down)
sub_eset  <- eSet[, which(eSet$Group %in% c("Control", "4NQO"))]
heatdata  <- t(scale(t(exprs(sub_eset)[top_genes, ])))
hc_row    <- hcopt(as.dist(1 - cor(t(heatdata))), method = "ward.D")
annot_col <- data.frame(Group = pData(sub_eset)$Group,
                         row.names = colnames(sub_eset))
pheatmap(heatdata,
          cluster_rows   = hc_row,
          show_rownames  = FALSE,
          annotation_col = annot_col,
          color          = colorRampPalette(c("blue", "white", "red"))(100),
          main           = "4NQO vs. Control DEGs")
```

### Save Results

```{r oscc-vs-ctrl-save, eval=FALSE}
saveRDS(list(up = oscc_ctrl_up, down = oscc_ctrl_down, full = res_oscc_ctrl),
         file.path(PATH, "results/deseq2_4nqo_vs_ctrl.RDS"))
```

## 4NQO + CPB vs. 4NQO {.tabset .tabset-fade .tabset-pills}

### DESeq2 Results

```{r cpb-vs-oscc-deseq2}
res_cpb_oscc <- run_deseq2(eSet,
                              groups    = c("4NQO", "4NQO_CPB"),
                              ref_group = "4NQO")
res_cpb_oscc <- res_cpb_oscc[!is.na(res_cpb_oscc$padj), ]
summary(res_cpb_oscc)

cat("\nSignificant DEGs (padj <= 0.05, |LFC| >= 1.5):\n")
print(c(
  "Up"   = sum(res_cpb_oscc$padj <= fdr & res_cpb_oscc$log2FoldChange >= lfc,  na.rm = TRUE),
  "Down" = sum(res_cpb_oscc$padj <= fdr & res_cpb_oscc$log2FoldChange <= -lfc, na.rm = TRUE)
))
```

### Volcano Plot

```{r cpb-vs-oscc-volcano}
plot_volcano(res_cpb_oscc, title = "4NQO + CPB vs. 4NQO")
```

### Top DEGs

```{r cpb-vs-oscc-top}
cpb_oscc_up   <- rownames(res_cpb_oscc[which(res_cpb_oscc$padj <= fdr &
                                                res_cpb_oscc$log2FoldChange >= lfc), ])
cpb_oscc_down <- rownames(res_cpb_oscc[which(res_cpb_oscc$padj <= fdr &
                                                res_cpb_oscc$log2FoldChange <= -lfc), ])

DT::datatable(as.data.frame(res_cpb_oscc[cpb_oscc_up, ]),
               caption = "4NQO + CPB vs. 4NQO - Upregulated")
DT::datatable(as.data.frame(res_cpb_oscc[cpb_oscc_down, ]),
               caption = "4NQO + CPB vs. 4NQO - Downregulated")
```

### Heatmap

```{r cpb-vs-oscc-heatmap}
top_genes <- c(cpb_oscc_up, cpb_oscc_down)
sub_eset  <- eSet[, which(eSet$Group %in% c("4NQO", "4NQO_CPB"))]
heatdata  <- t(scale(t(exprs(sub_eset)[top_genes, ])))
hc_row    <- hcopt(as.dist(1 - cor(t(heatdata))), method = "ward.D")
annot_col <- data.frame(Group = pData(sub_eset)$Group,
                         row.names = colnames(sub_eset))
pheatmap(heatdata,
          cluster_rows   = hc_row,
          show_rownames  = FALSE,
          annotation_col = annot_col,
          color          = colorRampPalette(c("blue", "white", "red"))(100),
          main           = "4NQO + CPB vs. 4NQO DEGs")
```

### Save Results

```{r cpb-vs-oscc-save, eval=FALSE}
saveRDS(list(up = cpb_oscc_up, down = cpb_oscc_down, full = res_cpb_oscc),
         file.path(PATH, "results/deseq2_4nqo_cpb_vs_4nqo.RDS"))
```

## 4NQO + CPB vs. Control {.tabset .tabset-fade .tabset-pills}

### DESeq2 Results

```{r cpb-vs-ctrl-deseq2}
res_cpb_ctrl <- run_deseq2(eSet,
                              groups    = c("Control", "4NQO_CPB"),
                              ref_group = "Control")
res_cpb_ctrl <- res_cpb_ctrl[!is.na(res_cpb_ctrl$padj), ]
summary(res_cpb_ctrl)

cat("\nSignificant DEGs (padj <= 0.05, |LFC| >= 1.5):\n")
print(c(
  "Up"   = sum(res_cpb_ctrl$padj <= fdr & res_cpb_ctrl$log2FoldChange >= lfc,  na.rm = TRUE),
  "Down" = sum(res_cpb_ctrl$padj <= fdr & res_cpb_ctrl$log2FoldChange <= -lfc, na.rm = TRUE)
))
```

### Volcano Plot

```{r cpb-vs-ctrl-volcano}
plot_volcano(res_cpb_ctrl, title = "4NQO + CPB vs. Control")
```

### Top DEGs

```{r cpb-vs-ctrl-top}
cpb_ctrl_up   <- rownames(res_cpb_ctrl[which(res_cpb_ctrl$padj <= fdr &
                                                res_cpb_ctrl$log2FoldChange >= lfc), ])
cpb_ctrl_down <- rownames(res_cpb_ctrl[which(res_cpb_ctrl$padj <= fdr &
                                                res_cpb_ctrl$log2FoldChange <= -lfc), ])

DT::datatable(as.data.frame(res_cpb_ctrl[cpb_ctrl_up, ]),
               caption = "4NQO + CPB vs. Control - Upregulated")
DT::datatable(as.data.frame(res_cpb_ctrl[cpb_ctrl_down, ]),
               caption = "4NQO + CPB vs. Control - Downregulated")
```

### Save Results

```{r cpb-vs-ctrl-save, eval=FALSE}
saveRDS(list(up = cpb_ctrl_up, down = cpb_ctrl_down, full = res_cpb_ctrl),
         file.path(PATH, "results/deseq2_4nqo_cpb_vs_ctrl.RDS"))
```

## DEG Overlap {.tabset}

### Venn Diagram

```{r venn}
library(VennDiagram)
library(grid)
venn_obj <- venn.diagram(
  x         = list(
    `4NQO vs Ctrl (up)`      = oscc_ctrl_up,
    `CPB vs 4NQO (up)`       = cpb_oscc_up,
    `CPB vs Ctrl (up)`       = cpb_ctrl_up
  ),
  filename  = NULL,
  fill      = c("steelblue", "tomato", "gold"),
  alpha     = 0.5,
  main      = "Overlap of Upregulated DEGs"
)
grid.newpage()
grid.draw(venn_obj)
```

## Session Info

```{r session-info}
sessionInfo()
```
