---
title: "4NQO OSCC Bcat - Bulk RNA-seq Pathway Enrichment Analysis"
author: "M. Muzamil Khan"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    theme: flatly
    toc: yes
    toc_float: true
  html_notebook:
    code_folding: hide
    theme: flatly
    toc: yes
    toc_float: true
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath("../.."))
```

```{r libraries, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(DT)
library(Biobase)
library(hypeR)
library(GSVA)
PATH <- getwd()
```

## Overview

This document performs overrepresentation analysis (ORA) using hypeR on differentially expressed genes and gene set variation analysis (GSVA) on the full expression matrix. Differential expression results are loaded from `02_diffanalysis.{Rmd, html}`.

Gene set collections used:

- **MSigDB Hallmarks** (H): curated gene sets representing well-defined biological states/processes
- **MSigDB C2 (KEGG/Reactome)**: curated canonical pathway gene sets
- **MSigDB C5 (GO)**: Gene Ontology gene sets
- **Wnt / beta-catenin signatures**: custom signatures for Wnt pathway activity

```{r load-data}
eSet         <- readRDS(file.path(PATH, "data/eSet_bulk.RDS"))
res_oscc_ctrl <- readRDS(file.path(PATH, "results/deseq2_4nqo_vs_ctrl.RDS"))
res_cpb_oscc  <- readRDS(file.path(PATH, "results/deseq2_4nqo_cpb_vs_4nqo.RDS"))
res_cpb_ctrl  <- readRDS(file.path(PATH, "results/deseq2_4nqo_cpb_vs_ctrl.RDS"))
```

```{r genesets}
# Load MSigDB gene sets (mouse via homolog mapping or direct mouse MSigDB)
msigdb_H  <- msigdb_gsets(species = "Mus musculus", category = "H")
msigdb_C2 <- msigdb_gsets(species = "Mus musculus", category = "C2",
                            subcategory = "CP:KEGG")
msigdb_C5 <- msigdb_gsets(species = "Mus musculus", category = "C5",
                            subcategory = "GO:BP")
```

## ORA: 4NQO vs. Control {.tabset .tabset-fade .tabset-pills}

### Hallmarks - Upregulated

```{r ora-oscc-ctrl-H-up}
hyp_oscc_ctrl_up_H <- hypeR(res_oscc_ctrl$up,
                               gsets    = msigdb_H,
                               test     = "hypergeometric",
                               background = nrow(eSet),
                               fdr      = 0.05,
                               quiet    = TRUE)
hyp_dots(hyp_oscc_ctrl_up_H, title = "4NQO vs. Control (Up) - Hallmarks")
```

### Hallmarks - Downregulated

```{r ora-oscc-ctrl-H-down}
hyp_oscc_ctrl_down_H <- hypeR(res_oscc_ctrl$down,
                                 gsets    = msigdb_H,
                                 test     = "hypergeometric",
                                 background = nrow(eSet),
                                 fdr      = 0.05,
                                 quiet    = TRUE)
hyp_dots(hyp_oscc_ctrl_down_H, title = "4NQO vs. Control (Down) - Hallmarks")
```

### KEGG - Upregulated

```{r ora-oscc-ctrl-C2-up}
hyp_oscc_ctrl_up_C2 <- hypeR(res_oscc_ctrl$up,
                               gsets    = msigdb_C2,
                               test     = "hypergeometric",
                               background = nrow(eSet),
                               fdr      = 0.05,
                               quiet    = TRUE)
hyp_dots(hyp_oscc_ctrl_up_C2, title = "4NQO vs. Control (Up) - KEGG")
```

### KEGG - Downregulated

```{r ora-oscc-ctrl-C2-down}
hyp_oscc_ctrl_down_C2 <- hypeR(res_oscc_ctrl$down,
                                  gsets    = msigdb_C2,
                                  test     = "hypergeometric",
                                  background = nrow(eSet),
                                  fdr      = 0.05,
                                  quiet    = TRUE)
hyp_dots(hyp_oscc_ctrl_down_C2, title = "4NQO vs. Control (Down) - KEGG")
```

## ORA: 4NQO + CPB vs. 4NQO {.tabset .tabset-fade .tabset-pills}

### Hallmarks - Upregulated

```{r ora-cpb-oscc-H-up}
hyp_cpb_oscc_up_H <- hypeR(res_cpb_oscc$up,
                              gsets    = msigdb_H,
                              test     = "hypergeometric",
                              background = nrow(eSet),
                              fdr      = 0.05,
                              quiet    = TRUE)
hyp_dots(hyp_cpb_oscc_up_H, title = "4NQO + CPB vs. 4NQO (Up) - Hallmarks")
```

### Hallmarks - Downregulated

```{r ora-cpb-oscc-H-down}
hyp_cpb_oscc_down_H <- hypeR(res_cpb_oscc$down,
                                gsets    = msigdb_H,
                                test     = "hypergeometric",
                                background = nrow(eSet),
                                fdr      = 0.05,
                                quiet    = TRUE)
hyp_dots(hyp_cpb_oscc_down_H, title = "4NQO + CPB vs. 4NQO (Down) - Hallmarks")
```

### KEGG - Upregulated

```{r ora-cpb-oscc-C2-up}
hyp_cpb_oscc_up_C2 <- hypeR(res_cpb_oscc$up,
                              gsets    = msigdb_C2,
                              test     = "hypergeometric",
                              background = nrow(eSet),
                              fdr      = 0.05,
                              quiet    = TRUE)
hyp_dots(hyp_cpb_oscc_up_C2, title = "4NQO + CPB vs. 4NQO (Up) - KEGG")
```

## GSVA Enrichment Analysis {.tabset .tabset-fade .tabset-pills}

### Hallmarks GSVA Scores

```{r gsva-H}
# Build gene set list for GSVA
gs_list_H <- lapply(msigdb_H$gsets, function(x) x$genes)
names(gs_list_H) <- names(msigdb_H$gsets)

gsva_scores_H <- gsva(exprs(eSet),
                       gs_list_H,
                       method    = "gsva",
                       kcdf      = "Poisson",
                       min.sz    = 5,
                       max.sz    = 500,
                       verbose   = FALSE)

saveRDS(gsva_scores_H, file.path(PATH, "results/gsva_hallmarks.RDS"))
```

### GSVA Heatmap

```{r gsva-H-heatmap}
library(pheatmap)
annot_col <- data.frame(Group = eSet$Group, row.names = colnames(eSet))
pheatmap(gsva_scores_H,
          annotation_col = annot_col,
          show_rownames  = TRUE,
          fontsize_row   = 6,
          color          = colorRampPalette(c("blue", "white", "red"))(100),
          main           = "GSVA - MSigDB Hallmarks")
```

### Wnt / Beta-catenin GSVA

```{r gsva-wnt}
# Custom Wnt/beta-catenin signatures (to be populated with published signatures)
wnt_sigs <- list(
  WNT_TARGET_GENES       = c("Axin2", "Lgr5", "Myc", "Ccnd1", "Cd44",
                               "Tcf7", "Lef1", "Tnfrsf19", "Nkd1", "Sp5"),
  CTNNB1_TARGETS_UP      = c("Ctnnb1", "Tcf4", "Axin2", "Dkk1", "Fzd1",
                               "Wnt3", "Wnt5a", "Sox17", "Fosl1", "Jun"),
  EMT_SIGNATURE          = c("Cdh2", "Vim", "Fn1", "Zeb1", "Zeb2",
                               "Twist1", "Snai1", "Snai2", "Mmp2", "Mmp9")
)

gsva_scores_wnt <- gsva(exprs(eSet),
                          wnt_sigs,
                          method  = "gsva",
                          kcdf    = "Poisson",
                          verbose = FALSE)

annot_col <- data.frame(Group = eSet$Group, row.names = colnames(eSet))
pheatmap(gsva_scores_wnt,
          annotation_col = annot_col,
          color          = colorRampPalette(c("blue", "white", "red"))(100),
          main           = "GSVA - Wnt/Beta-catenin Signatures")

saveRDS(gsva_scores_wnt, file.path(PATH, "results/gsva_wnt_signatures.RDS"))
```

## Session Info

```{r session-info}
sessionInfo()
```
